---
title: "Hands-on Exercise 04"
subtitle: "This exercise is a hands-on guide for Spatial Weights and Applications."
author: "Hung Thai"
date: "Sep 17, 2026"
date-modified: "last-modified"
execute:
  echo: true
  eval: true
  mesage: false
  warning: false
  freeze: true
format: 
  html:
    toc: true
    toc-depth: 4
    code-copy: true
editor: visual
---

# Spatio Weights and Application

## 1 Overview

This hands-on exercise will focus on how to compute spatial weight using R.

The targets of the exercise are:

- import geospatial data using appropriate function(s) of **sf** package,

- import csv file using appropriate function of **readr** package,

- perform relational join using appropriate join function of **dplyr** package,

- compute spatial weights using appropriate functions of **spdep** package, and

- calculate spatially lagged variables using appropriate functions of **spdep** package.

::: callout-note
### reference

[r4gdsa chap 08 - Spatial Weights and Applications](https://r4gdsa.netlify.app/chap08.html)[chap.](https://r4gdsa.netlify.app/chap06)
:::

## 2 The study area and data

Two data sets will be used in this hands-on exercise, they are:

- Hunan county boundary layer. This is a geospatial data set in ESRI shapefile format.

- *Hunan_2012.csv*: This csv file contains selected Hunan’s local development indicators in 2012.

## 3 Getting the Data and R Environment

Before import shapefile data, we need to install and loading the R packages

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse, knitr)
```

### 3.1 Importing the data into R environment.

Import Hunan shapefile in to R.

```{r}
hunan <- st_read(dsn = "../../data/Hands-on_Ex04/geospatial", 
                 layer = "Hunan")
```

Reading layer `Hunan' from data source`C:\hungthq\ISSS626-GAA\data\Hands-on_Ex04\geospatial' using driver \`ESRI Shapefile' Simple feature collection with 88 features and 7 fields\
Geometry type: POLYGON\
Dimension: XY\
Bounding box: xmin: 108.7831 ymin: 24.6342 xmax: 114.2544 ymax: 30.12812\
Geodetic CRS: WGS 84

### 3.2 Import csv file into r environment

```{r}
hunan2012 <- read_csv("../../data/Hands-on_Ex04/aspatial/Hunan_2012.csv")
```

```{r}
hunan2012 
```

### 3.2 Performing relational join

Joining the 2 table using the left_join() of dplyr package.

```{r}
hunan <- left_join(hunan,hunan2012) %>%
  select(1:4, 7, 15)
```

## 4 Visualising Regional Development Indicator

Preparing basemap and choropleth map showing the distribution of GDPPC 2012 by using qtm() of tmap package.

```{r}
basemap <- tm_shape(hunan) +
  tm_polygons() +
  tm_text("NAME_3", size=0.5)

gdppc <- qtm(hunan, "GDPPC") +
  tm_layout(
    legend.position = c("left", "bottom"))
tmap_arrange(basemap, gdppc, asp=1, ncol=2)
```

## 5 Computing contiguity spatial weights

Computing (QUEEN) contiguity based neighbours

```{r}
wm_q <- poly2nb(hunan, queen=TRUE)
summary(wm_q)
```

```         
Neighbour list object: 
Number of regions: 88  
Number of nonzero links: 448  
Percentage nonzero weights: 5.785124  
Average number of links: 5.090909  
Link number distribution:   
1  2  3  4  5  6  7  8  9 11   
2  2 12 16 24 14 11  4  2  1  
2 least connected regions: 
30 65 with 1 link 
1 most connected region: 
85 with 11 links
```

The summary report above shows that there are 88 area units in Hunan. The most connected area unit has 11 neighbours. There are two area units with only one heighbours.

For each polygon in our polygon object, *wm_q* lists all neighboring polygons. For example, to see the neighbors for the first polygon in the object, type:

```{r}
wm_q[[1]]
```

Polygon 1 has 5 neighbors. The numbers represent the polygon IDs as stored in hunan SpatialPolygonsDataFrame class.

We can retrive the county name of Polygon ID=1

```{r}
hunan$County[1]
```

Reveal the country names of the five neighbouring polygons:

```{r}
hunan$NAME_3[c(2,3,4,57,85)]
```

Retrieve the GDPPC of these five countries

```{r}
nb1 <- wm_q[[1]]
nb1 <- hunan$GDPPC[nb1]
nb1
```

Display the complete weight matrix by using str().

```{r}
str(wm_q)
```

### 5.1 Create (ROOK) contiguity based neighbours

```{r}
wm_r <- poly2nb(hunan, queen=FALSE)
summary(wm_r)
```

The summary report above shows that there are 88 area units in Hunan. The most connect area unit has 10 neighbours. There are two area units with only one heighbours.

### 5.2 Visualising contiguity weights

::: callout-important
To get our longitude values we map the st_centroid function over the geometry column of us.bound and access the longitude value through double bracket notation \[\[\]\] and 1. This allows us to get only the longitude, which is the first value in each centroid.
:::

```{r}
longitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[1]])
```

We do the same for latitude with one key difference. We access the second value per each centroid with \[\[2\]\].

```{r}
latitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[2]])
```

Put longitude and latitude into the same object

```{r}
coords <- cbind(longitude, latitude)
```

Check the first few observations to make sure it's correct.

```{r}
head(coords)
```

#### Plotting QUEEN contiguity based neighbours map

```{r}
plot(hunan$geometry, border="lightgrey")
plot(wm_q, coords, pch = 19, cex = 0.6, add = TRUE, col= "red")
```

#### Plotting Rook contiguity based neighbours map

```{r}
plot(hunan$geometry, border="lightgrey")
plot(wm_r, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

#### Plotting both Queen and Rook contiguity based neighbours maps

```{r}
par(mfrow=c(1,2))
plot(hunan$geometry, border="lightgrey", main="Queen Contiguity")
plot(wm_q, coords, pch = 19, cex = 0.6, add = TRUE, col= "red")
plot(hunan$geometry, border="lightgrey", main="Rook Contiguity")
plot(wm_r, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

## 6 Computing distance based neighbours

In this section, we will use dnearneigh() of spdap package to derive distance-based weight matrices.

The function identifies neighbours of region points by Euclidean distance with a distance band with lower d1= and upper d2= bounds controlled by the bounds= argument. If unprojected coordinates are used and either specified in the coordinates object x or with x as a two column matrix and longlat=TRUE, great circle distances in **km** will be calculated assuming the WGS84 reference ellipsoid.

### 6.1 Determine the cut-off distance

To determine the upper limit of distance band by using below steps:

- Return a matrix with the indices of points belonging to the set of the k nearest neighbours of each other by using [*knearneigh()*](https://r-spatial.github.io/spdep/reference/knearneigh.html) of **spdep**.

- Convert the knn object returned by *knearneigh()* into a neighbours list of class nb with a list of integer vectors containing neighbour region number ids by using [*knn2nb()*](https://r-spatial.github.io/spdep/reference/knn2nb.html).

- Return the length of neighbour relationship edges by using [*nbdists()*](https://r-spatial.github.io/spdep/reference/nbdists.html) of **spdep**. The function returns in the units of the coordinates if the coordinates are projected, in km otherwise.

- Remove the list structure of the returned object by using [**unlist()**](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/unlist).

```{r}
#coords <- coordinates(hunan)
k1 <- knn2nb(knearneigh(coords))
k1dists <- unlist(nbdists(k1, coords, longlat = TRUE))
summary(k1dists)
```

### 6.2 Computing fixed distance weight matrix

Compute the distance weight metrix

```{r}
wm_d62 <- dnearneigh(coords, 0, 62, longlat = TRUE)
wm_d62
```

::: callout-tip
## Quiz answer

Q: What is the meaning of “Average number of links: 3.681818”?

A:\
The **"Average number of links: 3.681818"** means that, on average across your spatial dataset, **each region has approximately 3.68 neighbors**.

Here is how that number is calculated and what it represents:

- **The Calculation:**

  \$\$\\text{Average Links} = \\frac{\\text{Total Nonzero Links}}{\\text{Total Regions}} = \\frac{324}{88} \\approx 3.681818\$\$
:::

Display the content of wm_d62 weight matrix

```{r}
str(wm_d62)
```

Alternative solution to display the structure of the weight matrix is to combine *table()* and *card()* of spdep.

```{r}
table(hunan$County, card(wm_d62))
```

```{r}
n_comp <- n.comp.nb(wm_d62)
n_comp$nc
```

```{r}
table(n_comp$comp.id)
```

### 6.3 Plotting fixed distance weight matrix

Plot the distance weight matrix

```{r}
plot(hunan$geometry, border="lightgrey")
plot(wm_d62, coords, add=TRUE)
plot(k1, coords, add=TRUE, col="red", length=0.08)
```

The red lines show the links of 1st nearest neighbours and the black lines show the links of neighbours within the cut-off distance of 62km.

Alternatively, we can plot both of them next to each other

```{r}
par(mfrow=c(1,2))
plot(hunan$geometry, border="lightgrey", main="1st nearest neighbours")
plot(k1, coords, add=TRUE, col="red", length=0.08)
plot(hunan$geometry, border="lightgrey", main="Distance link")
plot(wm_d62, coords, add=TRUE, pch = 19, cex = 0.6)
```

### 6.4 Computing adaptive distance weight matrix

One of the characteristics of fixed distance weight matrix is that more densely settled areas (usually the urban areas) tend to have more neighbours and the less densely settled areas (usually the rural counties) tend to have lesser neighbours. Having many neighbours smoothes the neighbour relationship across more neighbours.

It is possible to control the numbers of neighbours directly using k-nearest neighbours, either accepting asymmetric neighbours or imposing symmetry

```{r}
knn6 <- knn2nb(knearneigh(coords, k=6))
knn6
```

Display the content of the matrix by using *str()*.

```{r}
str(knn6)
```

#### Plotting distance based neighbours

```{r}
plot(hunan$geometry, border="lightgrey")
plot(knn6, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

## 7 Weights based on IDW

We will learn how to derive a spatial weight matrix based on Inversed Distance method in this section.

Compute the distances between areas by *nbdists()* of spdep.

```{r}
dist <- nbdists(wm_q, coords, longlat = TRUE)
ids <- lapply(dist, function(x) 1/(x))
ids
```

## 8 Row-standardised Weights Matrix

In this section, we need to assign weights to each neighboring polygon. In our case, each neighboring polygon will be assigned equal weight (style=“W”). This is accomplished by assigning the fraction 1/(#ofneighbors) to each neighboring county then summing the weighted income values. While this is the most intuitive way to summaries the neighbors’ values it has one drawback in that polygons along the edges of the study area will base their lagged values on fewer polygons thus potentially over- or under-estimating the true nature of the spatial autocorrelation in the data. For this example, we’ll stick with the style=“W” option for simplicity’s sake but note that other more robust options are available, notably style=“B”.

```{r}
rswm_q <- nb2listw(wm_q, style="W", zero.policy = TRUE)
rswm_q
```

The zero.policy=TRUE option allows for lists of non-neighbors. This should be used with caution since the user may not be aware of missing neighbors in their dataset however, a zero.policy of FALSE would return an error.

To see the weight of the first polygon’s eight neighbors type:

```{r}
rswm_q$weights[10]
```

Each neighbor is assigned a 0.125 of the total weight. This means that when R computes the average neighboring income values, each neighbor’s income will be multiplied by 0.125 before being tallied.

Using the same method, we can also derive a row standardised distance weight matrix

```{r}
rswm_ids <- nb2listw(wm_q, glist=ids, style="B", zero.policy=TRUE)
rswm_ids
```

```{r}
rswm_ids$weights[1]
```

```{r}
summary(unlist(rswm_ids$weights))
```

## 9 Application of spatial weight matrix

Learning points of this section:

- spatial lag with row-standardized weights,

- spatial lag as a sum of neighbouring values,

- spatial window average, and

- spatial window sum.

### 9.1 Spatial lag with row-standardized weights

Compute the average neighbor GDPPC value for each polygon

```{r}
GDPPC.lag <- lag.listw(rswm_q, hunan$GDPPC)
GDPPC.lag
```

Retrieved the GDPPC of these five countries

```{r}
nb1 <- wm_q[[1]]
nb1 <- hunan$GDPPC[nb1]
nb1
```

Append the spatially lag GDPPC values onto hunan sf data frame

```{r}
lag.list <- list(hunan$NAME_3, lag.listw(rswm_q, hunan$GDPPC))
lag.res <- as.data.frame(lag.list)
colnames(lag.res) <- c("NAME_3", "lag GDPPC")
hunan <- left_join(hunan,lag.res)
```

Display average neighboring income values (stored in the Inc.lag object) for each country

```{r}
head(hunan)
```

Now we plot both the GDPPC and spatial lag GDPPC for comparison.

```{r}
gdppc <- qtm(hunan, "GDPPC")
lag_gdppc <- qtm(hunan, "lag GDPPC")
tmap_arrange(gdppc, lag_gdppc, asp=1, ncol=2)
```

### 9.2 Spatial lag as a sum of neighboring values

We can calculate spatial lag as a sum of neighboring values by assigning binary weights. This requires us to go back to our neighbors list, then apply a function that will assign binary weights, then we use glist = in the nb2listw function to explicitly assign these weights.

We start by applying a function that will assign a value of 1 per each neighbor. This is done with lapply, which we have been using to manipulate the neighbors structure throughout the past notebooks.

```{r}
b_weights <- lapply(wm_q, function(x) 0*x + 1)
b_weights2 <- nb2listw(wm_q, 
                       glist = b_weights, 
                       style = "B")
b_weights2
```

With the proper weights assigned, we can use lag.listw to compute a lag variable from our weight and GDPPC.

```{r}
lag_sum <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
lag.res <- as.data.frame(lag_sum)
colnames(lag.res) <- c("NAME_3", "lag_sum GDPPC")
```

Examine the result:

```{r}
lag_sum
```

Now we will append the lag_sum GDPPC field into hunan sf data frame

```{r}
hunan <- left_join(hunan, lag.res)
```

Then we plot both the GDPPC and Spatial Lag Sum GDPPC for comparison

```{r}
gdppc <- qtm(hunan, "GDPPC")
lag_sum_gdppc <- qtm(hunan, "lag_sum GDPPC")
tmap_arrange(gdppc, lag_sum_gdppc, asp=1, ncol=2)
```

### 9.3 Spatial window average

The spatial window average uses row-standardized weights and includes the diagonal element. To do this in R, we need to go back to the neighbors structure and add the diagonal element before assigning weights.

To add the diagonal element to the neighbour list, we just need to use *include.self()* from spdep.

```{r}
wm_qs <- include.self(wm_q)
wm_qs
```

Take a good look at the neighbour list of area \[1\]

```{r}
wm_qs[[1]]
```

Notice that now \[1\] has six neighbours instead of five.

Now we obtain weights with *nb2listw()*

```{r}
wm_qs <- nb2listw(wm_qs)
wm_qs
```

Again, we use *nb2listw()* and *glist()* to explicitly assign weight values.

Then we create the lag variable from our weight structure and GDPPC variable.

```{r}
lag_w_avg_gpdpc <- lag.listw(wm_qs, 
                             hunan$GDPPC)
lag_w_avg_gpdpc
```

Next, we will convert the lag variable listw object into a data.frame

```{r}
lag.list.wm_qs <- list(hunan$NAME_3, lag.listw(wm_qs, hunan$GDPPC))
lag_wm_qs.res <- as.data.frame(lag.list.wm_qs)
colnames(lag_wm_qs.res) <- c("NAME_3", "lag_window_avg GDPPC")
```

::: callout-note
The third command line on the code chunk above renames the field names of *lag_wm_q1.res* object into *NAME_3* and *lag_window_avg GDPPC* respectively.
:::

We will append *lag_window_avg GDPPC* values onto *hunan* sf data.frame by using *left_join()* of **dplyr** package.

```{r}
hunan <- left_join(hunan, lag_wm_qs.res)
```

To compare the values of lag GDPPC and Spatial window average, `kable()` of Knitr package is used to prepare a table

```{r}
hunan %>%
  select("County", 
         "lag GDPPC", 
         "lag_window_avg GDPPC") %>%
  kable()
```

Then we plot the lag_gdppc and w_ave_gdppc maps next to each other for quick comparison.

```{r}
w_avg_gdppc <- qtm(hunan, "lag_window_avg GDPPC")
tmap_arrange(lag_gdppc, w_avg_gdppc, asp=1, ncol=2)
```

### 9.4 Spatial window sum

The spatial window sum is the counter part of the window average, but without using row-standardized weights.

To add the diagonal element to the neighbour list, we just need to use *include.self()* from **spdep**.

```{r}
wm_qs <- include.self(wm_q)
wm_qs
```

Next, we will assign binary weights to the neighbour structure that includes the diagonal element.

```{r}
b_weights <- lapply(wm_qs, function(x) 0*x + 1)
b_weights[1]
```

Again, we use *nb2listw()* and *glist()* to explicitly assign weight values.

```{r}
b_weights2 <- nb2listw(wm_qs, 
                       glist = b_weights, 
                       style = "B")
b_weights2
```

With our new weight structure, we can compute the lag variable with *lag.listw()*.

```{r}
w_sum_gdppc <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
w_sum_gdppc
```

Next, we will convert the lag variable listw object into a data.frame by using *as.data.frame()*.

```{r}
w_sum_gdppc.res <- as.data.frame(w_sum_gdppc)
colnames(w_sum_gdppc.res) <- c("NAME_3", "w_sum GDPPC")
```

::: callout-note
The second command line on the code chunk above renames the field names of *w_sum_gdppc.res* object into *NAME_3* and *w_sum GDPPC* respectively.
:::

Next, the code chunk below will be used to append *w_sum GDPPC* values onto *hunan* sf data.frame by using *left_join()* of **dplyr** package.

```{r}
hunan <- left_join(hunan, w_sum_gdppc.res)
```

To compare the values of lag GDPPC and Spatial window average, `kable()` of Knitr package is used to prepare a table.

```{r}
hunan %>%
  select("County", "lag_sum GDPPC", "w_sum GDPPC") %>%
  kable()
```

Lastly, *qtm()* of **tmap** package is used to plot the lag_sum GDPPC and w_sum_gdppc maps next to each other for quick comparison.

```{r}
w_sum_gdppc <- qtm(hunan, "w_sum GDPPC")
tmap_arrange(lag_sum_gdppc, w_sum_gdppc, asp=1, ncol=2)
```
