---
title: "Hands-on Exercise 04"
subtitle: "This exercise is a hands-on guide for Spatial Weights and Applications."
author: "Hung Thai"
date: "Sep 17, 2026"
date-modified: "last-modified"
execute:
  echo: true
  eval: true
  mesage: false
  warning: false
  freeze: true
format: 
  html:
    toc: true
    toc-depth: 4
    code-copy: true
editor: visual
---

# Spatio Weights and Application

## 1 Overview

This hands-on exercise will focus on how to compute spatial weight using R.

The targets of the exercise are:

- import geospatial data using appropriate function(s) of **sf** package,

- import csv file using appropriate function of **readr** package,

- perform relational join using appropriate join function of **dplyr** package,

- compute spatial weights using appropriate functions of **spdep** package, and

- calculate spatially lagged variables using appropriate functions of **spdep** package.

::: callout-note
### reference

[r4gdsa chap 08 - Spatial Weights and Applications](https://r4gdsa.netlify.app/chap08.html)[chap.](https://r4gdsa.netlify.app/chap06)
:::

## 2 The study area and data

Two data sets will be used in this hands-on exercise, they are:

- Hunan county boundary layer. This is a geospatial data set in ESRI shapefile format.

- *Hunan_2012.csv*: This csv file contains selected Hunan’s local development indicators in 2012.

## 3 Getting the Data and R Environment

Before import shapefile data, we need to install and loading the R packages

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse, knitr)
```

### Importing the data into R environment.

Import Hunan shapefile in to R.

```{r}
hunan <- st_read(dsn = "data/Hands-on_Ex04/geospatial", 
                 layer = "Hunan")
```

Reading layer `Hunan' from data source`C:\hungthq\ISSS626-GAA\data\Hands-on_Ex04\geospatial' using driver \`ESRI Shapefile' Simple feature collection with 88 features and 7 fields\
Geometry type: POLYGON\
Dimension: XY\
Bounding box: xmin: 108.7831 ymin: 24.6342 xmax: 114.2544 ymax: 30.12812\
Geodetic CRS: WGS 84

### Import csv file into r environment

```{r}
hunan2012 <- read_csv("data/Hands-on_Ex04/aspatial/Hunan_2012.csv")
```

```{r}
hunan2012 
```