---
title: "Hands-on Exercise 01"
subtitle: "In this exercise, we will load the data and do data wrangling."
author: "Hung Thai"
date: "Aug 24, 2026"
format: 
  html:
    toc: true
---

# Geospatial Data Science with R

[R for Geospatial Data Science and Analytics - 1  Geospatial Data Science with R](https://r4gdsa.netlify.app/chap01)

### Learning Outcome

- [x] installing and loading [sf](https://r-spatial.github.io/sf/) and [tidyverse](https://www.tidyverse.org/) packages into R environment,

- [ ] importing geospatial data by using appropriate functions of **sf** package,

- [ ] importing aspatial data by using appropriate function of **readr** package,

- [ ] exploring the content of simple feature data frame by using appropriate **Base R** and **sf** functions,

- [ ] assigning or transforming coordinate systems by using using appropriate *sf* functions,

- [ ] converting an aspatial data into a sf data frame by using appropriate function of **sf** package,

- [ ] performing geoprocessing tasks by using appropriate functions of **sf** package,

- [ ] performing data wrangling tasks by using appropriate functions of **dplyr** package and

- [ ] performing Exploratory Data Analysis (EDA) by using appropriate functions from **ggplot2** package.

### 1 The Data

| Dataset | Source | Description |
|------------------------|------------------------|------------------------|
| **Master Plan 2014 Subzone Boundary (Web)** | [data.gov.sg](https://data.gov.sg/) | Geospatial boundaries for Singapore's planning subzones. |
| **Pre-Schools Location** | [data.gov.sg](https://data.gov.sg/) | Location data for pre-schools in Singapore. |
| **Cycling Path** | [LTA DataMall](https://www.mytransport.sg/content/mytransport/home/dataMall.html) | Geospatial data for cycling paths in Singapore. |
| **Singapore Airbnb Listings** | [Inside Airbnb](http://insideairbnb.com/get-the-data.html) | Latest listing data for Airbnb properties in Singapore. |

### 2 Extracting the geospatial data sets

To get started, create a new folder called *Hands-on_Ex* in a working directory.

Next, create a new sub-folder called *Hands-on_ex01* in the newly created *Hands-on_Ex* folder.

In the `Hands-on_Ex01` folder, create a sub-folder called `data`. Then, inside the `data` sub-folder, create two sub-folders and name them `geospatial` and `aspatial` respectively.

Place `Master Plan 2014 Subzone Boundary (Web)`, `Pre-Schools Location` and `Cycling Path` zipped files into `geospatial` sub-folder and unzipped them. Copy the unzipped files from their respective sub-folders and place them inside `geospatial` sub-folder.

\*\*1.2.2 Extracting the aspatial data set\
\*\* Now, you will extract the downloaded `listing` data file. At `Downloads` folder, cut and paste `listing.csv` into `aspatial` sub-folder.

## 3 Getting Started

In this hands-on exercise, two R packages will be used. They are:

- **sf** for importing, managing, and processing geospatial data, and

- **tidyverse** for performing data science tasks such as importing, wrangling and visualising data.

Tidyverse consists of a family of R packages. In this hands-on exercise, the following packages will be used:

- **readr** for importing csv data,\
- **tidyr** for manipulating data,\
- **dplyr** for transforming data, and\
- **ggplot2** for visualising data

::: callout-note
Tidyverse consists of a family of R packages. In this hands-on exercise, the following packages will be used:

- **readr** for importing csv data,
- **tidyr** for manipulating data,
- **dplyr** for transforming data, and
- **ggplot2** for visualising data
:::

::: callout-warning
You must have **pacman** R package installed. If you have yet to do so, you should install **pacman** before moving to the next step.
:::

Use the code chunk below to load the necessary R packages into R.

```{r}
pacman::p_load(sf, tidyverse)
```

## 4 Importing Geospatial Data

### 4.1 Importing polygon feature data in shapefile format

The code chunk below uses *st_read()* function of **sf** package to import `MP14_SUBZONE_WEB_PL` shapefile into R as a polygon feature data frame. Note that when the input geospatial data is in shapefile format, two arguments will be used, namely: `dsn` to define the data path and `layer` to provide the shapefile name. Also note that no extension such as .shp, .dbf, .prj and .shx are needed.

```{r}
library(sf)
mpsz = st_read("data/geospatial/MasterPlan2014SubzoneBoundaryNoSea.geojson")
```

### 4.2 Importing polyline feature data in shapefile form

The code chunk below uses *st_read()* function of **sf** package to import `CyclingPath` shapefile into R as line feature data frame.

```{r}
cyclingpath = st_read("data/geospatial/CyclingPathNetworkGEOJSON.geojson")
```

### 4.3 Importing GIS data in kml format

The `PreSchoolsLocation` is in [**kml**](https://en.wikipedia.org/wiki/Keyhole_Markup_Language) format. The code chunk below will be used to import the kml into R.

```{r}
  preschool = st_read("data/geospatial/PreSchoolsLocation.geojson")

```

::: callout-tip
this is the example of a Tip
:::

## 5 Checking the Content

```{r}
  st_geometry(mpsz)
```

```{r}
glimpse(mpsz)
```

```{r}
head(mpsz, n=5)
```

## 6 Plotting the Geospatial Data

In geospatial data science, by looking at the feature information is not enough. We are also interested to visualise the geospatial features. This is the time you will find *plot()* of R Graphic comes in very handy as shown in the code chunk below.

```{r}
plot(mpsz)
```

```{r}
plot(st_geometry(mpsz))
```

```{r}
plot(mpsz["PLN_AREA_N"])
```

```{r}
plot(st_geometry(mpsz))
plot(st_geometry(preschool),
     add = TRUE)


```

## 7 Working with Projection

Map projection is an important property of a geospatial data. In order to perform geoprocessing using two geospatial data, we need to ensure that both geospatial data are projected using similar coordinate system.

In this section, you will learn how to project a simple feature data frame from one coordinate system to another coordinate system. The technical term of this process is called **projection transformation**.

### 7.1 Assigning EPSG code to a simple feature data frame

This is an example the coordinate system of `mpsz` simple feature data frame by using *st_crs()* of *sf* package as shown in the code chunk below.

```{r}
st_crs(mpsz)

```

Although `mpsz` data frame is projected in svy21 but when we read until the end of the print, it indicates that the EPSG is 9001. This is a wrong EPSG code because the correct EPSG code for svy21 should be [3414](https://epsg.io/3414).

In order to assign the correct EPSG code to `mpsz` data frame, *st_set_crs()* of **sf** package is used as shown in the code chunk below.

```{r}
mpsz <- st_set_crs(mpsz, 3414)
```

Now, let us check the CSR again by using the code chunk below.

```{r}
st_crs(mpsz)
```

### 7.2 Transforming the projection of preschool from wgs84 to svy21.

Let us perform the projection transformation preschool data by using the code chunk below.

```{r}
preschool <- st_transform(preschool, 
                          crs = 3414)
```

Now, let us try to plot the preschool layer ontop of mpsz layer again by using the similar code chunk you used earlier.

```{r}
preschool_svy21 <- st_transform(preschool, 3414)
mpsz <- st_set_crs(mpsz, 4326) %>% st_transform(3414)
#mpsz <- st_transform(mpsz, 3414)
```

```{r}
plot(st_geometry(mpsz))
plot(st_geometry(preschool_svy21), add = TRUE, col= "red", )
```

## 8. Importing and Converting An Aspatial Data

### 8.1 Importing the aspatial data

Since `listings` data set is in csv file format, we will use [*read_csv()*](https://readr.tidyverse.org/reference/read_delim.html) of **readr** package to import `listing.csv` as shown the code chunk below. The output R object is called `listings` and it is a [tibble data frame](https://r4ds.had.co.nz/tibbles.html).

```{r}
listings <- read_csv("data/aspatial/listings.csv")
```

After importing the data file into R, it is important for us to examine if the data file has been imported correctly.

The code chunk below shows *list()* of Base R instead of *glimpse()* is used to do the job.

```{r}
list(listings) 
```

### 8.2 Creating a simple feature data frame from an aspatial data frame

The code chunk below converts `listing` data frame into a simple feature data frame by using [*st_as_sf()*](https://r-spatial.github.io/sf/reference/st_as_sf.html) of **sf** packages

```{r}
library(dplyr)
listings_sf <- st_as_sf(listings, 
                        coords = c("longitude", "latitude"),
                        crs=4326) %>%
  st_transform(crs = 3414)
```

Examine the content of this newly created simple feature data frame.

```{r}
glimpse(listings_sf)
```

## 9 Geoprocessing with sf package

### 9.1 Use case 1: Land acquisition analysis

#### The scenario

The authority is planning to upgrade the exiting cycling path. To do so, they need to acquire 5 metres of reserved land on the both sides of the existing cycling path. You are tasked to determine the extend of the land need to be acquired and their total area.

#### The solution

Firstly, [`st_buffer()`](https://r-spatial.github.io/sf/reference/geos_unary.html) of **sf** package is used to compute the 5-meter buffers around cycling paths.

```{r}
library(sf)
library(dplyr)
buffer_cycling <- st_buffer(   cyclingpath, dist=5, nQuadSegs = 30)
```

This is followed by calculating the area of the buffers as shown in the code chunk below.

```{r}
buffer_cycling$AREA <- st_area(buffer_cycling)
```

If you are tidyverse person like me, code chunk below should be used.

```{r}
buffer_cycling <- buffer_cycling %>%   mutate(AREA = st_area(geometry))
```

```{r}
sum(buffer_cycling$AREA)
```

5317952 \[m\^2\]

We can also create a plot showing the buffer by a selected planning subzone.

Assuming that we are interested on the land acquisition in Tampines West planning subzone.

Firstly, `filter()` of dplyr package will be used to extract polygon feature of Tampines West by using the code chunk below.

```{r}
mpsz_selected <- mpsz %>%   filter(SUBZONE_N == "TAMPINES WEST") 
```

::: callout-note
You need to override the incorrect coordinate definition on `mpsz_selected` by explicitly setting it to WGS84 (EPSG 4326), and then transform it to the proper SVY21 metric system (EPSG 3414) to match your cycling data.
:::

Run this block of code to fix the projection:

```{r}
# 1. Force the correct WGS84 CRS definition onto the mislabeled subzone layer
mpsz_selected <- st_set_crs(mpsz_selected, 4326)

# 2. Transform the subzone layer coordinates to true SVY21 metres
mpsz_selected_svy21 <- st_transform(mpsz_selected, 3414)

# 3. Ensure your cycling layer is explicitly set to SVY21 as well
buffer_cycling <- st_transform(buffer_cycling, 3414)

# 4. Perform the intersection again
buffer_cycling_selected <- st_intersection(buffer_cycling, mpsz_selected_svy21)

# 5. Check the results
print(buffer_cycling_selected)
```

Plot the selected area

```{r}
#plot(st_geometry(buffer_cycling_selected))
```
