---
title: "Hands-on Exercise 01"
#author: "Hung Thai"
---

# **1  Geospatial Data Science with R**

## **1.1 Learning Outcome**

competent check boxes:

- [x] installing and loading [sf](https://r-spatial.github.io/sf/) and [tidyverse](https://www.tidyverse.org/) packages into R environment,

- [ ] importing geospatial data by using appropriate functions of **sf** package,

- [ ] importing aspatial data by using appropriate function of **readr** package,

- [ ] exploring the content of simple feature data frame by using appropriate **Base R** and **sf** functions,

- [ ] assigning or transforming coordinate systems by using using appropriate *sf* functions,

- [ ] converting an aspatial data into a sf data frame by using appropriate function of **sf** package,

- [ ] performing geoprocessing tasks by using appropriate functions of **sf** package,

- [ ] performing data wrangling tasks by using appropriate functions of **dplyr** package and

- [ ] performing Exploratory Data Analysis (EDA) by using appropriate functions from **ggplot2** package.

## **1.2 Data Acquisition**

1.2 Data Acquisition Data are key to data analytics including geospatial analytics. Hence, before analysing, we need to assemble the necessary data. In this hands-on exercise, you are required to extract the necessary data sets from the following sources:

- Master Plan 2014 Subzone Boundary (Web) from [data.gov.sg](https://data.gov.sg/)

- Pre-Schools Location from [data.gov.sg](https://data.gov.sg/)

- Cycling Path from [LTADataMall](https://www.mytransport.sg/content/mytransport/home/dataMall.html)

- Latest version of Singapore Airbnb listing data from [Inside Airbnb](http://insideairbnb.com/get-the-data.html)

### **1.2.1 Extracting the geospatial data sets**

To get started, create a new folder called *Hands-on_Ex* in a working directory.

Next, create a new sub-folder called *Hands-on_ex01* in the newly created *Hands-on_Ex* folder.

In the `Hands-on_Ex01` folder, create a sub-folder called `data`. Then, inside the `data` sub-folder, create two sub-folders and name them `geospatial` and `aspatial` respectively.

Place `Master Plan 2014 Subzone Boundary (Web)`, `Pre-Schools Location` and `Cycling Path` zipped files into `geospatial` sub-folder and unzipped them. Copy the unzipped files from their respective sub-folders and place them inside `geospatial` sub-folder.

\*\*1.2.2 Extracting the aspatial data set\
\*\* Now, you will extract the downloaded `listing` data file. At `Downloads` folder, cut and paste `listing.csv` into `aspatial` sub-folder.

## **1.3 Getting Started**

In this hands-on exercise, two R packages will be used. They are:

- **sf** for importing, managing, and processing geospatial data, and

- **tidyverse** for performing data science tasks such as importing, wrangling and visualising data.

Tidyverse consists of a family of R packages. In this hands-on exercise, the following packages will be used:

- **readr** for importing csv data,\
- **tidyr** for manipulating data,\
- **dplyr** for transforming data, and\
- **ggplot2** for visualising data

::: callout-note
Tidyverse consists of a family of R packages. In this hands-on exercise, the following packages will be used:

- **readr** for importing csv data,
- **tidyr** for manipulating data,
- **dplyr** for transforming data, and
- **ggplot2** for visualising data
:::

::: callout-warning
You must have **pacman** R package installed. If you have yet to do so, you should install **pacman** before moving to the next step.
:::

Use the code chunk below to load the necessary R packages into R.

```{r}
pacman::p_load(sf, tidyverse)
```

## **1.4 Importing Geospatial Data**

### **1.4.1 Importing polygon feature data in shapefile format**

The code chunk below uses *st_read()* function of **sf** package to import `MP14_SUBZONE_WEB_PL` shapefile into R as a polygon feature data frame. Note that when the input geospatial data is in shapefile format, two arguments will be used, namely: `dsn` to define the data path and `layer` to provide the shapefile name. Also note that no extension such as .shp, .dbf, .prj and .shx are needed.

```{r}
library(sf)
mpsz = st_read("data/geospatial/MasterPlan2014SubzoneBoundaryNoSea.geojson")
```

### **1.4.2 Importing polyline feature data in shapefile form**

The code chunk below uses *st_read()* function of **sf** package to import `CyclingPath` shapefile into R as line feature data frame.

```{r}
cyclingpath = st_read("data/geospatial/CyclingPathNetworkGEOJSON.geojson")
```

### **1.4.3 Importing GIS data in kml format**

The `PreSchoolsLocation` is in [**kml**](https://en.wikipedia.org/wiki/Keyhole_Markup_Language) format. The code chunk below will be used to import the kml into R.

```{r}
  preschool = st_read("data/geospatial/PreSchoolsLocation.geojson")

```

::: callout-tip
this is the example of a Tip
:::

## **1.5 Checking the Content**

```{r}
  st_geometry(mpsz)
```

```{r}
glimpse(mpsz)
```

```{r}
head(mpsz, n=5)
```

## **1.6 Plotting the Geospatial Data**

In geospatial data science, by looking at the feature information is not enough. We are also interested to visualise the geospatial features. This is the time you will find *plot()* of R Graphic comes in very handy as shown in the code chunk below.

```{r}
plot(mpsz)
```

```{r}
plot(st_geometry(mpsz))
```

```{r}
plot(mpsz["PLN_AREA_N"])
```

```{r}
plot(st_geometry(mpsz))
plot(st_geometry(preschool),
     add = TRUE)


```
