---
title: "Hands-on Exercise 06"
subtitle: "Geographical Segmentation"
author: "Hung Thai"
date: "Oct 01, 2026"
date-modified: "last-modified"
execute:
  echo: true
  eval: true
  mesage: false
  warning: false
  freeze: true
format: 
  html:
    toc: true
    toc-depth: 4
    code-copy: true
editor: visual
---

# 1. Overview

This exercise provides practical experience in delineating homogeneous regions from multivariate spatial data using two key methods: hierarchical clustering and spatially constrained clustering.

In this practical session, we will learn how to:

1.  Import and convert GIS polygons to `sf` data frames and `SpatialPolygonsDataFrame` objects.

2.  Run hierarchical clustering with `hclust()`.

3.  Implement spatially constrained clustering with `spdep::skater()`.

4.  Create thematic maps and diagnostic plots using `tmap` and `ggplot2`.

## The Analytical Question

Regional delineation using multivariate spatial data is a core task in market analysis and spatial policy. Here, we aim to partition Shan State, Myanmar, into homogeneous regions using six household ICT ownership indicators: **Radio, Television, Landline phone, Mobile phone, Computer,** and **Home internet access**.

## The data

Two data sets will be used in this study. They are:

- Myanmar Township Boundary Data (i.e. *myanmar_township_boundaries*) : This is a GIS data in ESRI shapefile format. It consists of township boundary information of Myanmar. The spatial data are captured in polygon features.

- *Shan-ICT.csv*: This is an extract of [**The 2014 Myanmar Population and Housing Census Myanmar**](https://myanmar.unfpa.org/en/publications/2014-population-and-housing-census-myanmar-data-sheet) at the township level.

Both data sets are download from [Myanmar Information Management Unit (MIMU)](http://themimu.info/)

```{r}
#| echo: false
pacman::p_load(spdep, tmap, sf, ClustGeo, 
               ggpubr, cluster, factoextra, NbClust,
               heatmaply, corrplot, psych, tidyverse, GGally)
```

# 2. Data Import and Preparation

## Import Myanmar Township Boundary GIS data and its associated attrbiute table

```{r}
shan_sf <- st_read(dsn = "data/geospatial", 
                   layer = "myanmar_township_boundaries") %>%
  filter(ST %in% c("Shan (East)", "Shan (North)", "Shan (South)")) %>%
  select(c(2:7))
```

View the content of the newly created shan_sf simple features data

```{r}
shan_sf
```

Reveal the data type of it’s fields

```{r}
glimpse(shan_sf)
```

## Importing aspatial data

Import InfoComm data

```{r}
ict <- read_csv ("data/aspatial/Shan-ICT.csv")
```

summary statistics of ict data.frame

```{r}
summary(ict)
```

## Derive new variables using dplyr package

Because the raw variables represent total counts of households, using them directly introduces size bias. Townships with larger overall populations will naturally record higher numbers of households owning radios, TVs, and other devices. To control for population scale, we normalize each indicator by calculating its household penetration rate.

```{r}
ict_derived <- ict %>%
  mutate(`RADIO_PR` = `Radio`/`Total households`*1000) %>%
  mutate(`TV_PR` = `Television`/`Total households`*1000) %>%
  mutate(`LLPHONE_PR` = `Land line phone`/`Total households`*1000) %>%
  mutate(`MPHONE_PR` = `Mobile phone`/`Total households`*1000) %>%
  mutate(`COMPUTER_PR` = `Computer`/`Total households`*1000) %>%
  mutate(`INTERNET_PR` = `Internet at home`/`Total households`*1000) %>%
  rename(`DT_PCODE` =`District Pcode`,`DT`=`District Name`,
         `TS_PCODE`=`Township Pcode`, `TS`=`Township Name`,
         `TT_HOUSEHOLDS`=`Total households`,
         `RADIO`=`Radio`, `TV`=`Television`, 
         `LLPHONE`=`Land line phone`, `MPHONE`=`Mobile phone`,
         `COMPUTER`=`Computer`, `INTERNET`=`Internet at home`) 
```

Review the summary statistics of the newly derived penetration rates

```{r}
summary(ict_derived)
```

# 3. Exploratory Data Analysis (EDA)

## 3.1 EDA using statistical graphics

Using histogram to identify the overall distribution of the data values

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

Using Boxplot to detect if there are outliers.

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_boxplot(color="black", 
               fill="light blue")
```

Plotting the distribution of the newly derived variables (i.e. Radio penetration rate)

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO_PR`)) +
  geom_boxplot(color="black", 
               fill="light blue")
```

Multiple histograms are plotted to reveal the distribution of the selected variables in the ict_derived data.frame

```{r}
#| echo: false
radio <- ggplot(data=ict_derived, 
                aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

tv <- ggplot(data=ict_derived, 
             aes(x= `TV_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

llphone <- ggplot(data=ict_derived, 
                  aes(x= `LLPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

mphone <- ggplot(data=ict_derived, 
                 aes(x= `MPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

computer <- ggplot(data=ict_derived, 
                   aes(x= `COMPUTER_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

internet <- ggplot(data=ict_derived, 
                   aes(x= `INTERNET_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")


```

```{r}
ggarrange(radio, tv, llphone, mphone, computer, internet, 
          ncol = 3, 
          nrow = 2)
```

## 3.2 EDA using choropleth map

### 3.2.1 Joining geospatial data with aspatial data

Before constructing the choropleth map, we must join the spatial dataset (shan_sf) with the aspatial attribute table (ict_derived). We will use dplyr’s left_join(), keeping shan_sf as the base simple feature data frame and appending ict_derived as the join table using the unique identifier TS_PCODE. The code chunk below executes this operation.

```{r}
shan_sf <- left_join(shan_sf, 
                     ict_derived, by=c("TS_PCODE"="TS_PCODE"))

write_rds(shan_sf, "data/rds/shan_sf.rds")
```

The data fields from ict_derived data frame are now updated into the data frame of shan_sf.

```{r}
shan_sf <- read_rds("data/rds/shan_sf.rds")
```

### 3.2.2 Preparing a choropleth map

Use choropleth map to have a quick look at the distribution of Radio penetration rate of Shan State at township level

```{r}
# qtm(shan_sf, "RADIO_PR")
#| fig-width: 12
#| fig-height: 8
tm_shape(shan_sf) +
  tm_polygons(fill = "RADIO_PR") +
  tm_title("Radio density per region", size = 1.5) +
  tm_layout(legend.position = c("right", "bottom"),
            legend.text.size = 0.4,
            legend.title.size = 0.5)
```

::: callout-note
Be noted that the original script from prof is not working due to the data field name is not correct ('RADIO_PR' instead of 'RADIO_PR.y'). We need to verify the field name of the data and correct it accordingly.
:::

```{r}
#| echo: false
names(shan_sf)
```

To demonstrate that the spatial patterns in the previous choropleth map are biased by township population size, we construct two comparative choropleth maps using the code chunk below: one displaying the total number of households (TT_HOUSEHOLDS.map) and the other showing households owning radios (RADIO.map).

```{r}
#| fig-width: 16
#| fig-height: 8
TT_HOUSEHOLDS.map <- tm_shape(shan_sf) + 
  tm_fill(col = "TT_HOUSEHOLDS",
          n = 5,
          style = "jenks", 
          title = "Total households") + 
  tm_borders(alpha = 0.5) + 
  tm_title("Myanmar Township total Households", size = 2.5) +
  tm_layout(legend.position = c("right", "bottom"),
            legend.text.size = 0.7,
            legend.title.size = 1)

RADIO.map <- tm_shape(shan_sf) + 
  tm_fill(col = "RADIO",
          n = 5,
          style = "jenks",
          title = "Number Radio ") + 
  tm_borders(alpha = 0.5) +
  tm_title("Myanmar Township number of radios", size = 2.5) +
  tm_layout(legend.position = c("right", "bottom"),
            legend.text.size = 0.7,
            legend.title.size = 1)

tmap_arrange(TT_HOUSEHOLDS.map, RADIO.map,
             asp=NA, ncol=2)
```

Notice that the choropleth maps above confirm this pattern: townships with larger household populations naturally exhibit higher absolute numbers of radio owners. To control for this population scale effect, we now plot side-by-side choropleth maps comparing the total number of households against the actual radio penetration rate.

```{r}
#| fig-width: 16
#| fig-height: 8

tm_shape(shan_sf) +
  tm_polygons(c("TT_HOUSEHOLDS", "RADIO_PR"),
              style="jenks") +
  tm_facets(sync = TRUE, ncol = 2) +
  tm_legend(legend.position = c("right", "bottom"))+
  #tm_layout(outer.margins=0, asp=0) +
  tm_layout(legend.position = c("right", "bottom"),
          legend.text.size = 0.7,
          legend.title.size = 1)

```

We can see that there are some regions that the radio own rate are extremely high despite of the number of house holds in the township is quite low.

# 4. Correlation Analysis

Before performing cluster analysis, it is essential to ensure that the input variables are not highly correlated, as multicollinearity can disproportionately weight overlapping dimensions. In this section, you will learn how to use the corrplot.mixed() function from the corrplot package to visualize and evaluate the correlation structure among the variables.

```{r}
cluster_vars.cor = cor(ict_derived[,12:17])
corrplot.mixed(cluster_vars.cor,
               lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black")

```

The correlation plot above shows that COMPUTER_PR and INTERNET_PR are highly correlated. This suggest that only one of them should be used in the cluster analysis instead of both.

# 5. Hierarchy Cluster Analysis

## 5.1 Extracting clustering variables

Extract the clustering variables from the shan_sf simple feature object into data.frame.

```{r}
cluster_vars <- shan_sf %>%
  st_set_geometry(NULL) %>%
  select("TS.x", "RADIO_PR", "TV_PR", "LLPHONE_PR", "MPHONE_PR", "COMPUTER_PR")
head(cluster_vars,10)
```

Notice that INTERNET_PR.y has been excluded from the final clustering variables due to its high correlation with COMPUTER_PR.y, preventing redundancy in the distance matrix. Next, we set the township names as row identifiers instead of relying on default row indices.

```{r}
row.names(cluster_vars) <- cluster_vars$"TS.x"
head(cluster_vars,10)
```

```{r}
shan_ict <- select(cluster_vars, c(2:6))
head(shan_ict, 10)
```

Save to rds file

```{r}
write_rds(shan_ict, "data/rds/shan_ict.rds")
```

::: callout-note
original script from prof: shan_ict \<- read_rds("chap12/data/rds/shan_ict.rds")\
\
I assume this script is to save, not to read so i changed the script to save_rds
:::

## 5.2 Data Standardisation

Because cluster analysis typically incorporates multiple variables measured on disparate scales, variables with larger magnitudes or wider ranges can disproportionately dominate distance calculations. To prevent this bias, it is standard practice to standardise the input variables before running the cluster analysis.

## 5.3 Min-Max standardisation

In the code chunk below, the normalize() function from the heatmaply package is used to standardise the clustering variables via the Min-Max scaling method. The summary() function is then called to inspect the summary statistics of the rescaled variables.

```{r}
shan_ict.std <- normalize(shan_ict)
summary(shan_ict.std)
```

## 5.4 Z-score standadisation

Using scale() of Base R to perform Z-score standardization.

```{r}
shan_ict.z <- scale(shan_ict)
describe(shan_ict.z)
```

::: callout-warning
Z-score standardisation method should only be used if we would assume all variables come from some normal distribution.
:::

## 5.5 Visualising the standardised clustering variables

Beyond reviewing numerical summary statistics, it is good practice to inspect the distributions of the standardised clustering variables graphically. The code chunk below visualises the distribution of the scaled RADIO_PR variable.

```         
```

```{r}
#| eval: false
#| include: false

#r <- 
ggplot(data=ict_derived, 
            aes(x= "RADIO_PR")) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Raw values without standardisation")


shan_ict_s_df <- as.data.frame(shan_ict.std)
#s <- 
ggplot(data=shan_ict_s_df, 
            aes(x="RADIO_PR")) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Min-Max Standardisation")

shan_ict_z_df <- as.data.frame(shan_ict.z)
#z <- 
ggplot(data=shan_ict_z_df, 
            aes(x="RADIO_PR")) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Z-score Standardisation")


#ggarrange(r, s, z,
#          ncol = 2,
#         nrow = 1)

```

```{r}
#| eval: false
#| include: false
r <- ggplot(data=ict_derived, 
            aes(x= `RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Raw values without standardisation")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s <- ggplot(data=shan_ict_s_df, 
            aes(x=`RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Min-Max Standardisation")

shan_ict_z_df <- as.data.frame(shan_ict.z)
z <- ggplot(data=shan_ict_z_df, 
            aes(x=`RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Z-score Standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)

```

## 5.6 Computing proximity matrix

While multiple R packages provide functions to calculate distance matrices, we will compute the proximity matrix using Base R’s dist(). The function supports six distance metrics: Euclidean, maximum, Manhattan, Canberra, binary, and Minkowski, with Euclidean distance used as the default.

```{r}
proxmat <- dist(shan_ict, method = 'euclidean')
```

list the content of proxmat for visual inspection.

```{r}
proxmat
```

## 5.7 Computing hierarchical clustering

Several R packages provide hierarchical clustering routines, but in this hands-on exercise, we use hclust() from the core stats package. The function applies an agglomerative approach and supports eight linkage algorithms: ward.D, ward.D2, single, complete, average (UPGMA), mcquitty (WPGMA), median (WPGMC), and centroid (UPGMC).

The code chunk below performs hierarchical cluster analysis using the ward.D method. The resulting output is stored as an hclust object, which represents the hierarchical tree structure produced by the clustering process.

```{r}
hclust_ward <- hclust(proxmat, method = 'ward.D')
```

plot the tree by using plot() of R Graphics

```{r}
#| fig-width: 14
#| fig-height: 8
plot(hclust_ward, cex = 1)
```

## 5.8 Selecting the optimal clustering algorithm

A key challenge in hierarchical cluster analysis is identifying the linkage method that yields the strongest clustering structure. This can be assessed using the agnes() function from the cluster package. Operating similarly to hclust(), agnes() additionally provides the agglomerative coefficient ($AC$), which measures the strength of the clustering structure on a scale from 0 to 1 (with values closer to 1 indicating well-defined clusters).The code chunk below calculates and compares the agglomerative coefficients across several hierarchical clustering algorithms to identify the best-performing method.

```{r}
m <- c( "average", "single", "complete", "ward")
names(m) <- c( "average", "single", "complete", "ward")

ac <- function(x) {
  agnes(shan_ict, method = x)$ac
}

map_dbl(m, ac)
```

With reference to the output above, we can see that Ward’s method provides the strongest clustering structure among the four methods assessed. Hence, in the subsequent analysis, only Ward’s method will be used.

## 5.9 Determining Optimal Clusters

Another key technical challenge in cluster analysis is determining the optimal number of clusters to retain. Three standard heuristic and statistical methods are commonly used to guide this decision:

- **Elbow Method** (Within-cluster sum of squares)

- **Average Silhouette Method**

- **Gap Statistic Method**

### 5.9.1 Gap Statistic Method

The Gap Statistic compares the total intra-cluster variation across different values of $k$ with their expected values under a null reference distribution generated from a uniform random baseline. The estimated optimal number of clusters is the value of $k$ that maximizes the gap statistic, indicating that the observed cluster structure departs most significantly from random noise. To compute this metric, we use the clusGap() function from the cluster package.

```{r}
set.seed(12345)
gap_stat <- clusGap(shan_ict, 
                    FUN = hcut, 
                    nstart = 25, 
                    K.max = 10, 
                    B = 50)
# Print the result
print(gap_stat, method = "firstmax")
```

Note that the hcut() function is supplied by the factoextra package to serve as the hierarchical clustering wrapper. Next, we can visualise the resulting gap statistic curve using factoextra's fviz_gap_stat() function to identify the optimal number of clusters.

```{r}
fviz_gap_stat(gap_stat)
```

Based on the gap statistic plot above, the default criterion suggests $k = 1$. However, retaining a single cluster is practically uninformative for regional segmentation. Upon closer inspection of the curve, $k = 6$ produces the highest local peak and represents the most viable alternative cluster solution.Note: Beyond these standard diagnostic curves, the NbClust package (Charrad et al., 2014) provides 30 distinct validation indices. It automates consensus voting across combinations of cluster numbers, distance measures, and clustering algorithms to recommend an optimal grouping scheme.

### 5.9.2 Elbow Method

The Elbow Method plots the total within-cluster sum of squares (WSS) against $k$. We look for the "bend" or "elbow" where the rate of decline flattens.

```{r}
elbow <- fviz_nbclust(shan_ict, 
                      FUNcluster = hcut, 
                      method = "silhouette",
                      hc_method = "ward.D",
                      k.max = 10)

```

```{r}
elbow
```

### 5.9.3 Average Silhouette Method

The Average Silhouette Method measures how well each observation lies within its cluster. A high average silhouette width indicates good clustering quality. The dashed line automatically identifies the optimal $k$ that maximizes the metric.

```{r}
fviz_nbclust(shan_ict, 
             FUNcluster = hcut, 
             method = "silhouette",
             hc_method = "ward.D",
             k.max = 10) 
```

## 5.10 Interpreting the dendrograms

In the dendrogram above, each leaf represents an individual observation (township). Moving up the tree, similar observations are agglomerated into branches, which iteratively fuse at greater heights. The vertical axis represents the fusion height, which reflects the degree of dissimilarity between clusters—the greater the height, the less similar the merged groups are. Crucially, the proximity of two observations can only be inferred from the height at which their branches first intersect; their horizontal ordering along the x-axis provides no indication of similarity. To visually demarcate the selected clusters, we can superimpose bounding boxes using the rect.hclust() function from R's stats package, specifying outline colors via the border argument.

```{r}
#| fig-width: 14
#| fig-height: 8
plot(hclust_ward, cex = 1)
rect.hclust(hclust_ward, 
            k = 6, 
            border = 2:5)
```

## 5.11 Visually-driven hierarchical clustering analysis

In this section, we explore visually driven hierarchical cluster analysis using the heatmaply package. heatmaply enables the creation of both rich interactive heatmaps and publication-ready static displays.

### 5.11.1 Transforming the Data Frame into a Matrix 

Because clustering heatmap functions require numeric matrix inputs rather than standard data frames, we must first convert the data structure. The code chunk below transforms the shan_ict data frame into a numeric data matrix.

```{r}
shan_ict_mat <- data.matrix(shan_ict)
```

### 5.11.2 Plotting interactive cluster heatmap using heatmaply()

```{r}
heatmaply(normalize(shan_ict_mat),
          Colv=NA,
          dist_method = "euclidean",
          hclust_method = "ward.D",
          seriate = "OLO",
          colors = Blues,
          k_row = 6,
          margins = c(NA,200,60,NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main="Geographic Segmentation of Shan State by ICT indicators",
          xlab = "ICT Indicators",
          ylab = "Townships of Shan State"
)

```

## 5.12 Mapping the clusters formed

Closer inspection of the dendrogram above supports the selection of a six-cluster solution. In the code chunk below, Base R's `cutree()` function is used to prune the tree and extract cluster memberships for this six-cluster model.

```{r}
groups <- as.factor(cutree(hclust_ward, k=6))
```

The output vector, `groups`, contains the cluster membership assignments. To map and analyze these clusters spatially, they must be joined to the `shan_sf` simple feature object.

The code chunk below performs this integration across three steps:

1.  Convert the `groups` vector into a matrix.

2.  Use `cbind()` to append the cluster matrix to `shan_sf`, producing a new simple feature object named `shan_sf_cluster`.

3.  Use `rename()` from the `dplyr` package to rename the newly appended column (`as.matrix.groups.`) to `CLUSTER`.

```{r}
shan_sf_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER`=`as.matrix.groups.`)
```

Next, *qtm()* of **tmap** package is used to plot the choropleth map showing the cluster formed.

```{r}
#| fig-height: 10
#| fig-width: 8

qtm(shan_sf_cluster, "CLUSTER") +
  tm_legend(legend.position = c("right", "bottom"))+
  #tm_layout(outer.margins=0, asp=0) +
  tm_layout(legend.position = c("right", "bottom"),
            legend.text.size = 0.7,
            legend.title.size = 1)
      

```

# 6. Spatially Constrained Clustering: SKATER approach

## 6.1 Converting into SpatialPolygonsDataFrame

First, we must convert shan_sf into a SpatialPolygonsDataFrame. This step is necessary because the skater() function in the spdep package requires legacy sp objects rather than modern sf simple features. In the code chunk below, we use as_Spatial() from the sf package to coerce shan_sf into a SpatialPolygonsDataFrame named shan_sp.

```{r}
shan_sp <- as_Spatial(shan_sf)
```

## 6.2 Computing Neighbour List

```{r}
shan.nb <- poly2nb(shan_sp)
summary(shan.nb)
```

We can visualise the spatial adjacency graph across Shan State using the code chunk below. To provide geographic context, the connectivity network is overlaid onto the township boundary polygons. The first `plot()` call renders the base polygon boundaries of `shan_sp`. The subsequent `plot()` call draws the neighbour list object (`nb`), using township polygon centroids as network nodes. By specifying `col = "blue"` and setting `add = TRUE`, the neighbour connectivity graph is drawn directly atop the boundary layer.

```{r}
coords <- st_coordinates(
  st_centroid(st_geometry(shan_sf)))
```

```{r}
#| fig-width: 8
#| fig-height: 8
plot(st_geometry(shan_sf), 
     border=grey(.5))
plot(shan.nb,
     coords, 
     col="blue", 
     add=TRUE)
```

Note that rendering the connectivity network before the polygon boundaries causes spatial clipping. Base R graphics define the canvas bounding box (xlim and ylim) based entirely on the initial plot() call. Because the boundary polygons cover a larger spatial extent than the centroid network nodes, plotting the boundary layer first ensures the entire geographic region is displayed without truncation.

## 6.3 Computing minimum spanning tree

**Calculating edge costs**

Next, [*nbcosts()*](https://r-spatial.github.io/spdep/reference/nbcosts.html) of **spdep** package is used to compute the cost of each edge. It is the distance between it nodes. This function compute this distance using a data.frame with observations vector in each node.

```{r}
lcosts <- nbcosts(shan.nb, shan_ict)
```

hese values quantify the multivariate dissimilarity (cost) between each township and its contiguous neighbours across the five variables, functioning as generalized spatial weights.

Using `spdep::nb2listw()`, we convert the neighbour list into a spatial weights object by assigning the calculated `lcosts` as edge weights. Setting `style = "B"` ensures a binary, non-row-standardized structure so the raw dissimilarity costs are preserved.

```{r}
shan.w <- nb2listw(shan.nb, 
                   lcosts, 
                   style="B")
summary(shan.w)
```

## 6.4 Computing minimum spanning tree

The minimum spanning tree is computed by mean of the [*mstree()*](https://r-spatial.github.io/spdep/reference/mstree.html) of **spdep** package as shown in the code chunk below.

```{r}
shan.mst <- mstree(shan.w)
```

After computing the MST, we can check its class and dimension by using the code chunk below.

```{r}
class(shan.mst)
```

```{r}
dim(shan.mst)
```

Note that the dimension is 54 and not 55. This is because the minimum spanning tree consists on n-1 edges (links) in order to traverse all the nodes.

We can display the content of *shan.mst* by using *head()* as shown in the code chunk below.

```{r}
head(shan.mst)
```

The plot method for the MST include a way to show the observation numbers of the nodes in addition to the edge. As before, we plot this together with the township boundaries. We can see how the initial neighbour list is simplified to just one edge connecting each of the nodes, while passing through all the nodes.

```{r}
#| fig-width: 8
#| fig-height: 8
plot(st_geometry(shan_sf), 
     border=gray(.5))
plot.mst(shan.mst, 
         coords, 
         col="blue", 
         cex.lab=0.7, 
         cex.circles=0.005, 
         add=TRUE)
```

## 6.5 Computing spatially constrained clusters using SKATER method

compute the spatially constrained cluster using [*skater()*](https://r-spatial.github.io/spdep/reference/skater.html) of **spdep** package.

```{r}
clust6 <- spdep::skater(edges = shan.mst[,1:2], 
                        data = shan_ict, 
                        method = "euclidean", 
                        ncuts = 5)
```

The *skater()* takes three mandatory arguments: - the first two columns of the MST matrix (i.e. not the cost), - the data matrix (to update the costs as units are being grouped), and - the number of cuts. Note: It is set to **one less than the number of clusters**. So, the value specified is **not** the number of clusters, but the number of cuts in the graph, one less than the number of clusters.

The result of the *skater()* is an object of class **skater**. We can examine its contents by using the code chunk below.

```{r}
str(clust6)
```

The most interesting component of this list structure is the groups vector containing the labels of the cluster to which each observation belongs (as before, the label itself is arbitary). This is followed by a detailed summary for each of the clusters in the edges.groups list. Sum of squares measures are given as ssto for the total and ssw to show the effect of each of the cuts on the overall criterion.

We can check the cluster assignment by using the conde chunk below.

```{r}
ccs6 <- clust6$groups
ccs6
```

We can find out how many observations are in each cluster by means of the table command. Parenthetially, we can also find this as the dimension of each vector in the lists contained in edges.groups. For example, the first list has node with dimension 12, which is also the number of observations in the first cluster.

```{r}
table(ccs6)
```

Lastly, we can also plot the pruned tree that shows the five clusters on top of the townshop area.

```{r}
#| fig-width: 8
#| fig-height: 8

plot(st_geometry(shan_sf), 
     border=gray(.5))
plot(clust6, 
     coords, 
     cex.lab=.7,
     groups.colors=c("red","green","blue", "brown", "pink"),
     cex.circles=0.005, 
     add=TRUE)

```

## 6.6 Visualising the clusters in choropleth map

plot the newly derived clusters by using SKATER method.

```{r}
#| fig-width: 8
#| fig-height: 8

groups_mat <- as.matrix(clust6$groups)
shan_sf_spatialcluster <- cbind(shan_sf_cluster, as.factor(groups_mat)) %>%
  rename(`SP_CLUSTER`=`as.factor.groups_mat.`)
qtm(shan_sf_spatialcluster, "SP_CLUSTER") +
  tm_legend(legend.position = c("right", "bottom"))+
  tm_layout(legend.position = c("right", "bottom"),
            legend.text.size = 0.7,
            legend.title.size = 1)
```

For easy comparison, it will be better to place both the hierarchical clustering and spatially constrained hierarchical clustering maps next to each other.

```{r}
#| fig-width: 14
#| fig-height: 8
hclust.map <- qtm(shan_sf_cluster,
                  "CLUSTER") + 
  tm_borders(alpha = 0.5) + 
  tm_title("hierarchical clustering", size = 2) +
  tm_legend(legend.position = c("right", "bottom"))+
  tm_layout(legend.position = c("right", "bottom"),
            legend.text.size = 0.6,
            legend.title.size = 1) 
  

shclust.map <- qtm(shan_sf_spatialcluster,
                   "SP_CLUSTER") + 
  tm_borders(alpha = 0.5) +
  tm_title("spatially constrained hierarchical clustering", size = 2) +
  tm_legend(legend.position = c("right", "bottom"))+
  tm_layout(legend.position = c("right", "bottom"),
            legend.text.size = 0.6,
            legend.title.size = 1)

tmap_arrange(hclust.map, shclust.map,
             asp=NA, ncol=2)
```

# 7. Spatially Constrained Clustering: ClustGeo Method

In this section, you will gain hands-on experience on using functions provided by **ClustGeo** package to perform non-spatially constrained hierarchical cluster analysis and spatially constrained cluster analysis.

## 7.1 About ClustGeo package

The `ClustGeo` package is specifically designed for spatially constrained cluster analysis. In particular, its `hclustgeo()` function implements a Ward-like hierarchical clustering algorithm that incorporates geographical contiguity constraints alongside attribute data.

The algorithm balances two dissimilarity matrices—$D_0$ (attribute dissimilarity) and $D_1$ (spatial distance/constraint matrix)—using a mixing parameter, $\alpha \in [0, 1]$. At each agglomeration step, the objective criterion minimizes a convex combination of attribute and spatial homogeneity. The goal is to select an $\alpha$ value that enhances spatial coherence without significantly compromising cluster quality in the attribute space. To identify this trade-off, `ClustGeo` provides the `choicealpha()` diagnostic function.

## 7.2 Ward-like hierarchical clustering: ClustGeo

ClustGeo package provides function called `hclustgeo()` to perform a typical Ward-like hierarchical clustering just like `hclust()` you learned in previous section.

To perform non-spatially constrained hierarchical clustering, we only need to provide the function a dissimilarity matrix as shown in the code chunk below.

```{r}
nongeo_cluster <- hclustgeo(proxmat)
plot(nongeo_cluster, cex = 0.5)
rect.hclust(nongeo_cluster, 
            k = 6, 
            border = 2:5)
```

**Mapping the clusters formed**

Similarly, we can plot the clusters on a categorical area shaded map by using the steps we learned in 5.7.12 Mapping the clusters formed.

```{r}
groups <- as.factor(cutree(nongeo_cluster, k=6))
```

```{r}
shan_sf_ngeo_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)
```

```{r}
#| fig-width: 8
#| fig-height: 8
qtm(shan_sf_ngeo_cluster, "CLUSTER") +
  tm_borders(alpha = 0.5) + 
  tm_title("hierarchical clustering", size = 2) +
  tm_legend(legend.position = c("right", "bottom"))+
  tm_layout(legend.position = c("right", "bottom"),
            legend.text.size = 0.6,
            legend.title.size = 1) 
```

## 7.3 Spatially Constrained Hierarchical Clustering

Before we can performed spatially constrained hierarchical clustering, a spatial distance matrix will be derived by using [`st_distance()`](https://r-spatial.github.io/sf/reference/geos_measures.html) of sf package.

```{r}
dist <- st_distance(shan_sf, shan_sf)
distmat <- as.dist(dist)
```

Notice that `as.dist()` is used to convert the data frame into matrix.

Next, `choicealpha()` will be used to determine a suitable value for the mixing parameter alpha as shown in the code chunk below.

```{r}
cr <- choicealpha(proxmat, distmat, range.alpha = seq(0, 1, 0.1), K=6, graph = TRUE)
```

With reference to the graphs above, alpha = 0.2 will be used as shown in the code chunk below.

```{r}
clustG <- hclustgeo(proxmat, distmat, alpha = 0.2)
```

Next, `cutree()` is used to derive the cluster objecct.

```{r}
groups <- as.factor(cutree(clustG, k=6))
```

We will then join back the group list with *shan_sf* polygon feature data frame by using the code chunk below.

```{r}
shan_sf_Gcluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)
```

We can now plot the map of the newly delineated spatially constrained clusters.

```{r}
#| fig-width: 8
#| fig-height: 8
qtm(shan_sf_Gcluster, "CLUSTER") +
  tm_borders(alpha = 0.5) + 
  tm_title("hierarchical G-clustering", size = 2) +
  tm_legend(legend.position = c("right", "bottom"))+
  tm_layout(legend.position = c("right", "bottom"),
            legend.text.size = 0.6,
            legend.title.size = 1) 
```

# 8. Visual Interpretation of Clusters

## 8.1 Visualising individual clustering variable

Code chunk below is used to reveal the distribution of a clustering variable (i.e RADIO_PR) by cluster.

```{r}
ggplot(data = shan_sf_ngeo_cluster,
       aes(x = CLUSTER, y = RADIO_PR)) +
  geom_boxplot()
```

The boxplot reveals Cluster 3 displays the highest mean Radio Ownership Per Thousand Household. This is followed by Cluster 2, 1, 4, 6 and 5.

## 8.2 Multivariate Visualisation

Past studies shown that parallel coordinate plot can be used to reveal clustering variables by cluster very effectively. In the code chunk below, [`ggparcoord()`](https://ggobi.github.io/ggally/reference/ggparcoord.html) of [**GGally**](https://ggobi.github.io/ggally/index.html) package

```{r}
ggparcoord(data = shan_sf_ngeo_cluster, 
           columns = c(17:21), 
           scale = "globalminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of ICT Variables by Cluster") +
  facet_grid(~ CLUSTER) + 
  theme(axis.text.x = element_text(angle = 30))
```

The parallel coordinates plot above reveals distinct socioeconomic profiles across the groups: households in Cluster 4 townships exhibit the highest ownership rates of televisions and mobile phones, whereas Cluster 5 exhibits the lowest adoption rates across all five ICT indicators.

When visualizing high-dimensional data, the `scale` argument in `GGally::ggparcoord()` provides several transformation methods to normalize the clustering variables:

- **`std`:** Standardizes variables univariately by subtracting the mean and dividing by the standard deviation.

- **`robust`:** Applies robust univariate scaling by centering on the median and dividing by the median absolute deviation (MAD).

- **`uniminmax`:** Rescales each variable univariately to a uniform range between $0$ and $1$.

- **`globalminmax`:** Preserves raw values without transformation, framing coordinates against global extrema.

- **`center`:** Applies `uniminmax` scaling and centers each variable at a statistic specified by `scaleSummary`.

- **`centerObs`:** Applies `uniminmax` scaling and centers coordinates around a baseline observation designated by `centerObsID`.

Because no single scaling method fits every analytical scenario, selecting the appropriate transformation depends on your data distribution and interpretative objectives. Finally, numerical summaries—such as the mean, median, and standard deviation—should accompany visual plots to corroborate profile findings. In the code chunk below, we use `group_by()` and `summarise()` from `dplyr` to compute cluster-level mean values for each ICT variable.

```{r}
shan_sf_ngeo_cluster %>% 
  st_set_geometry(NULL) %>%
  group_by(CLUSTER) %>%
  summarise(mean_RADIO_PR = mean(RADIO_PR),
            mean_TV_PR = mean(TV_PR),
            mean_LLPHONE_PR = mean(LLPHONE_PR),
            mean_MPHONE_PR = mean(MPHONE_PR),
            mean_COMPUTER_PR = mean(COMPUTER_PR))
```
