---
title: "Hands-on Exercise 02"
subtitle: "This exercise is a hands-on guide for using the tmap package in R to create both static and interactive choropleth maps."
author: "Hung Thai"
date: "Sep 01, 2026"
format: 
  html:
    toc: true
    toc-depth: 4
editor: visual
---

# Spatial Point Patterns Analysis

## Overview

Spatial Point Pattern Analysis is the evaluation of the pattern, or distribution, of a set of points on a surface. It can refer to the actual spatial or temporal location of these points or also include data from point sources. It is one of the most fundamental concepts in geography and spatial analysis.

This exercise has 2 parts:

**Part1 covers First-order Spatial Point Patterns Analysis Methods** which focus on understanding the intensity or density of points across a study area. It examines how the distribution of points varies over space, essentially identifying trends or patterns in point density. This type of analysis deals with the individual locations of points and their distribution, without considering interactions between them.

**Part2 cover Second-order Spatial Point Patterns Analysis** Methods which examine the spatial relationships between points in a pattern, specifically focusing on how the presence of one point influences the location of others. It goes beyond simply describing the overall density of points (first-order effects) by investigating clustering, dispersion, or randomness at various spatial scales.

::: callout-note
## references

[r4gdsa chap 04 - First-order Spatial Point Patterns Analysis Methods](https://r4gdsa.netlify.app/chap04.html)

[r4gdsa chap 05 - Second-order Spatial Point Patterns Analysis Methods](https://r4gdsa.netlify.app/chap05)
:::

## Part 1 First-order Spatial Point Pattern Analysis methods

In this part, [spatstat](https://cran.r-project.org/web/packages/spatstat/) is used to perform commonly used 1st-PPA methods in order to answer the specific questions:

- Are the children centres in Singapore randomly distributed throughout the country?

- If the answer is not, then the next logical question is where are the locations with higher concentration of childcare centres?

Clark-Even Test for Nearest Neighbour Analysis Method & Kernel Density Estimation Method will be used to answer the above questions.

### The data

To provide answers to the questions above, two data sets will be used. They are:

- [*Child Care Services*](https://data.gov.sg/datasets?query=Child+Care+Services&resultId=d_5d668e3f544335f8028f546827b773b4) data from data.gov.sg, a point feature data providing both location and attribute information of childcare centres.

- [*Master Plan 2019 Subzone Boundary (No Sea*)](https://data.gov.sg/datasets?agencies=Urban+Redevelopment+Authority+(URA)&resultId=d_8594ae9ff96d0c708bc2af633048edfb), a polygon feature data providing information of URA 2019 Master Plan Planning Subzone boundary data.

### Installing and Loading the R packages

```{r}
pacman::p_load(sf, terra, spatstat, tmap, rvest, tidyverse)
```

### Importing and wrangling geospatial data sets

Import and transform Master Plan 2019 Subzone (No Sea) data set into R environment.

```{r}
mpsz_sf <- st_read("../../data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaGEOJSON.geojson") |> st_zm(drop = TRUE, what = "ZM") |> st_transform(crs = 3414)
```

```         
head(mpsz_sf, n=3) 
```

```{r, results='markup'}
head(mpsz_sf, n=3)
```

Clean up island areas

```{r}
mpsz_cl <- mpsz_sf %>%
  filter(SUBZONE_N != "SOUTHERN GROUP",
         PLN_AREA_N != "WESTERN ISLANDS",
         PLN_AREA_N != "NORTH-EASTERN ISLANDS")
```

```{r}
head(mpsz_cl, n=3)
```

Import and transform the childcare Service data downloaded from data.gov.sg into R environment

```{r}
childcare_sf <- st_read("../../data/geospatial/ChildCareServices.geojson") |>
  st_zm(drop = TRUE, what = "ZM") |>
  st_transform(crs = 3414)
```

### Mapping the geospatial data sets

```{r}
plot(st_geometry(mpsz_cl))
plot(st_geometry(childcare_sf), 
     add = TRUE)
```

Alternatively, we can also using tmap to create an interactive point symbol map

```{r}
tmap_mode('view')
tm_shape(childcare_sf)+
  tm_dots()
```

::: callout-important
Running `tmap_mode("plot")` resets the plotting engine to generate standard static maps instead of interactive Leaflet maps (`tmap_mode("view")`), ensuring subsequent chunks do not consume extra web connections.
:::

```{r}
tmap_mode('plot')
```

::: callout-tip
Prof Kam mentioned in the note: "Currently, three internet map layers are provided. They are: ESRI.WorldGrayCanvan, OpenStreetMap, and ESRI.WorldTopoMap. The default is ESRI.WorldGrayCanvas.".\
\
In order to try new map, we can change the R code to different map as below:

```{r}
#Change default map to OpenStreetMap
tmap_mode("view")

tm_basemap("OpenStreetMap") +
  tm_shape(childcare_sf) +
  tm_dots()

tmap_mode("plot")

```

we also can add options to choose the world map in map layers:

```{r}

tmap_mode("view")
#add 3 map types in to the basemap
tmap_options(basemap.server = c("Esri.WorldTopoMap", "OpenStreetMap", "Esri.WorldGrayCanvas"))

tm_shape(childcare_sf)+
  tm_dots()

tmap_mode("plot")
```
:::

### Geospatial data wrangling

**Converting sf data frames to ppp class**

Using as.ppp() of **spatstat** package to convert *childcare_sf* to ppp format

```{r}
childcare_ppp <- as.ppp(childcare_sf)
```

Next, class() of Base R will be used to verify the object class of *childcare_ppp*.

```{r}
class(childcare_ppp)
```

summary statistics of the newly converted ppp

```{r}
summary(childcare_ppp)
```

**Creating *owin* object**

***owin*** (Observation Window) is an object class that defines the exact geographical boundary or spatial domain within which point events are observed and analyzed.

Point pattern analysis requires a known spatial boundary to accurately compute metrics such as intensity (points per unit area), edge-correction factors, and nearest neighbor distances.

```{r}
sg_owin <- as.owin(mpsz_cl)
```

```{r}
class(sg_owin)
```

Display sg_owin object

```{r}
plot(sg_owin)
```

**Combining point events object and owin object**

Extract childcare events located within Singapore.

```{r}
childcareSG_ppp = childcare_ppp[sg_owin]
```

```{r}
childcareSG_ppp
```

### Clark-Evan Test for Nearest Neighbour Analysis

::: callout-note
## Info:

**Nearest Neighbor Analysis (NNA)** is a spatial statistics method that calculates the average distance between each point and its closest neighbor to determine if a pattern of points is clustered, dispersed, or randomly distributed.
:::

Perform the Clark-Evans test of aggregation for a spatial point pattern by using [*clarkevans.test()*](https://cran.r-project.org/web/packages/spatstat.explore/refman/spatstat.explore.html#clarkevans.test) of [**spatstat.explore**](https://cran.r-project.org/web/packages/spatstat.explore/refman/spatstat.explore.html) package.

The test hypotheses are:

Ho = The distribution of childcare services are randomly distributed.

H1= The distribution of childcare services are not randomly distributed.

The 95% confident interval will be used.

**Perform the Clark-Evan test without CSR**

```{r}
clarkevans.test(childcareSG_ppp,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"))
```

::: callout-important
## Quiz answer:

- **Decision:** Since the p-value ($< 2.2 \times 10^{-16}$) is well below the significance threshold of $\alpha = 0.05$, we reject the null hypothesis $H_0$.

- **Inference:** With an aggregation index of $R = 0.53532$ ($R < 1$), there is statistically significant evidence at the 95% confidence level to conclude that childcare centres in Singapore are spatially clustered rather than randomly distributed.
:::

::: callout-important
## Business communication

- **Market Clustering & Co-location:** Childcare facilities are heavily concentrated in specific geographical clusters rather than dispersed evenly across Singapore. This concentration typically aligns with densely populated residential estates (such as major HDB towns) and transit-accessible hubs where young families reside.

- **Competitive Density vs. Underserved Gaps:** High-density zones indicate strong existing demand but also heightened commercial competition among operators. Conversely, areas with sparse coverage may reveal untapped market potential or underserved gaps requiring targeted facility planning or government support.
:::

**Perform the Clark-Evans test with CSR**

Clark-Evans test with CSR using MonteCarlo method.

```{r}
clarkevans.test(childcareSG_ppp,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"),
                method="MonteCarlo",
                nsim=99)
```

::: callout-important
## Statistical conclusion & communication

- **Decision:** Since the Monte Carlo p-value ($p = 0.01$) is less than the significance threshold $\alpha = 0.05$, we reject the null hypothesis $H_0$.

- **Inference:** With an aggregation index of $R = 0.53532$ ($R < 1$), there is statistically significant evidence at the 95% confidence level (based on 99 CSR simulations) that childcare centres in Singapore exhibit a **spatially clustered** pattern rather than random spatial dispersion.

**Business communication:**

Childcare centres are not uniformly scattered across Singapore; they naturally cluster around high-density demand nodes such as major residential towns (e.g., mature HDB estates) and key public transit interchanges where young working families live and commute.
:::

### Kernel Density Estimation Method

::: callout-note
## Info

**Kernel Density Estimation (KDE)** is a valuable tool for visualising and analyzing first-order spatial point patterns. It is widely considered a method within **Exploratory Spatial Data Analysis (ESDA)** because it’s used to visualize and understand spatial data patterns by transforms discrete point data (like locations of childcare service, crime incidents or disease cases) into continuous density surfaces that reveal clusters and variations in event occurrences, without making prior assumptions about data distribution. It helps to begin understanding data distribution, identify hotspots, and explore relationships between spatial variables before performing more rigorous analysis.
:::

**Working with automatic bandwidth selection method**

Compute a kernel density by using *density()* of **spatstat**:

```{r}
kde_SG_diggle <- density(
  childcareSG_ppp,
  sigma=bw.diggle,
  edge=TRUE,
  kernel="gaussian") 
```

```{r}
plot(kde_SG_diggle)
```

```{r}
summary(kde_SG_diggle)
```

```{r}
bw <- bw.diggle(childcareSG_ppp)
bw
```

**Rescalling KDE values**

Convert the unit of measurement from meter to kilometer.

```{r}
childcareSG_ppp_km <- rescale.ppp(
  childcareSG_ppp, 1000, "km")
```

re-run `density()` using the resale data set and plot the output kde map.

```{r}
kde_childcareSG_km <- density(childcareSG_ppp_km,
                              sigma=bw.diggle,
                              edge=TRUE,
                              kernel="gaussian")
```

```{r}
plot(kde_childcareSG_km)
```

Now we can see the data values changed.

**Working with different automatic bandwidth methods**

::: callout-tip
## automatic bandwidth methods in spatstat

+-----------------------+----------------------------------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------+
|                       |                                                                                                                                  |                                                                                  |
+=======================+==================================================================================================================================+==================================================================================+
| **Function**          | **Method / Description**                                                                                                         | **Best Used For**                                                                |
+-----------------------+----------------------------------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------+
| `bw.diggle()`         | **Cross-validation (Diggle's method):** Minimizes the mean squared error (MSE) between the estimated and true intensity.         | Detecting a single tight cluster amidst random noise.                            |
+-----------------------+----------------------------------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------+
| `bw.ppl()`            | **Likelihood Cross-Validation:** Selects bandwidth by maximizing the leave-one-out cross-validated log-likelihood.               | Patterns predominantly composed of tight clusters.                               |
+-----------------------+----------------------------------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------+
| `bw.CvL()`            | **Cronie and van Lieshout’s method:** Based on the Campbell formula to select bandwidth.                                         | Point patterns where points are closely packed or general spatial trends.        |
+-----------------------+----------------------------------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------+
| `bw.scott()`          | **Scott's Rule of Thumb:** Computes separate bandwidths for the x and y dimensions based on standard deviations and sample size. | Fast parametric approximation across anisotropic (directional) spatial patterns. |
+-----------------------+----------------------------------------------------------------------------------------------------------------------------------+----------------------------------------------------------------------------------+
:::

```{r}
bw.CvL(childcareSG_ppp_km)
```

```{r}
bw.scott(childcareSG_ppp_km)
```

```{r}
bw.ppl(childcareSG_ppp_km)
```

```{r}
bw.diggle(childcareSG_ppp_km)
```

Comparing output of using *bw.diggle* and *bw.ppl* methods:

```{r}
kde_childcareSG.ppl <- density(childcareSG_ppp_km, 
                               sigma=bw.ppl, 
                               edge=TRUE,
                               kernel="gaussian")
par(mfrow=c(1,2))
plot(kde_childcareSG_km, main = "bw.diggle")
plot(kde_childcareSG.ppl, main = "bw.ppl")
```

**Working with different kernel methods**

::: callout-tip
## Kernel methods used in density.ppp()

In **`spatstat`**, the kernel smoothing function (passed via the `kernel` argument in *`density.ppp()`*) defines the weighting shape used to spread each point's mass across the surface.

**Available Kernel Methods**

- **`"gaussian"` (Default):** A bell-shaped normal distribution curve. It produces the smoothest surface and extends infinitely (truncated computationally), making it the standard choice for most point pattern analyses.

- **`"epanechnikov"`:** A inverted parabolic kernel ($1 - r^2$). It is mathematically optimal in minimizing asymptotic mean integrated squared error (MISE) and has bounded, finite support.

- **`"quartic"` (or Biweight):** A smooth, bell-like polynomial function with finite support ($(1 - r^2)^2$). It drops off more steeply at the boundaries than Gaussian while remaining smoother than Epanechnikov.

- **`"disc"`:** A flat, uniform cylindrical kernel where all points within the bandwidth radius receive equal weight, resulting in sharp, step-like density transitions.
:::

```{r, fig.width=10, fig.height=8}
par(mfrow=c(2,2))
plot(density(childcareSG_ppp_km, 
             sigma=0.2959712, 
             edge=TRUE, 
             kernel="gaussian"), 
     main="Gaussian")
plot(density(childcareSG_ppp_km, 
             sigma=0.2959712, 
             edge=TRUE, 
             kernel="epanechnikov"), 
     main="Epanechnikov")
plot(density(childcareSG_ppp_km, 
             sigma=0.2959712, 
             edge=TRUE, 
             kernel="quartic"), 
     main="Quartic")
plot(density(childcareSG_ppp_km, 
             sigma=0.2959712, 
             edge=TRUE, 
             kernel="disc"), 
     main="Disc")
```

::: callout-tip
I have error: "Error in plot.new(): figure margins too large" when plotting the above figures.\
This error occurs when the R plotting pane/device is physically too small to fit the requested layout (multi-panel grids & legends).\
\
Quick fix:

Define bigger width & height:\
\`\`\`{r, fig.width=10, fig.height=8}
:::

### Fixed and Adaptive KDE

**Computing KDE by using fixed bandwidth**

In this sample, we will compute a KDE layer by defining a bandwidth of 600 meter.

```{r}
kde_childcareSG_fb <- density(childcareSG_ppp_km,
                              sigma=0.6, 
                              edge=TRUE,
                              kernel="gaussian")
plot(kde_childcareSG_fb)
```

**Computing KDE by using adaptive bandwidth**

Derive adaptive kernel density estimation by using [*density.adaptive()*](https://rdrr.io/cran/spatstat/man/adaptive.density.html) of **spatstat**.

```{r}
kde_childcareSG_ab <- adaptive.density(
  childcareSG_ppp_km, 
  method="kernel")
plot(kde_childcareSG_ab)
```

Compare the fixed and adaptive kernel density estimation outputs

```{r}
par(mfrow=c(1,2))
plot(kde_childcareSG_fb, main = "Fixed bandwidth")
plot(kde_childcareSG_ab, main = "Adaptive bandwidth")
```

::: callout-note
+-------------------------------------------------+-----------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------+
|                                                 |                                                                                                           |                                                                                                                  |
+=================================================+===========================================================================================================+==================================================================================================================+
| **Feature**                                     | **Fixed Bandwidth KDE**                                                                                   | **Adaptive Bandwidth KDE**                                                                                       |
+-------------------------------------------------+-----------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------+
| **Bandwidth Radius (**$\sigma$)                 | **Constant** across the entire study region. Every point uses the exact same search radius (e.g., 600 m). | **Variable** across space. The radius adjusts dynamically based on local point density.                          |
+-------------------------------------------------+-----------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------+
| **High-Density Areas (e.g., Urban hubs)**       | May oversmooth distinct local clusters and merge separate peaks together.                                 | Narrows the bandwidth (smaller $\sigma$) to capture sharper, local hotspot peaks.                                |
+-------------------------------------------------+-----------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------+
| **Low-Density Areas (e.g., Rural / Periphery)** | Leaves excessive empty space or creates artificial, isolated noise spikes around solitary points.         | Widens the bandwidth (larger $\sigma$) to smooth out noise and provide meaningful generalized trends.            |
+-------------------------------------------------+-----------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------+
| **Best Used For**                               | Homogeneous point distributions or when comparing absolute density across regions with standard units.    | Highly skewed distributions where point densities vary drastically across urban and rural/peripheral landscapes. |
+-------------------------------------------------+-----------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------+
| **Implementation in `spatstat`**                | `density(childcare_ppp, sigma = 0.6)`                                                                     | `adaptive.density(childcare_ppp, method = "kernel")`                                                             |
+-------------------------------------------------+-----------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------+
:::

### Plotting cartographic quality KDE map

**Converting gridded output into raster**

Using *rast()* of **terra** package to convert the im kernel density objects into SpatRaster objects.

```{r}
kde_childcareSG_bw_terra <- rast(kde_childcareSG_km)
```

```{r}
class(kde_childcareSG_bw_terra)
```

```{r}
kde_childcareSG_bw_terra
```

**Assigning projection systems**

Using *crs()* of **terra** to assign the CRS information on *kde_childcareSG_bw_terra* layer.

```{r}
crs(kde_childcareSG_bw_terra) <- "EPSG:3414"
```

```{r}
kde_childcareSG_bw_terra
```

```         
coord. ref. : SVY21 / Singapore TM (EPSG:3414)
```

**Plotting KDE map with tmap**

Display the raster in cartographic quality map using **tmap** package.

```{r}
tm_shape(kde_childcareSG_bw_terra) + 
  tm_raster(col.scale = 
              tm_scale_continuous(
                values = "viridis"),
            col.legend = tm_legend(
              title = "Density values",
              title.size = 0.7,
              text.size = 0.7,
              bg.color = "white",
              bg.alpha = 0.7,
              position = tm_pos_in(
                "right", "bottom"),
              frame = TRUE)) +
  tm_graticules(labels.size = 0.7) +
  tm_compass() +
  tm_layout(scale = 1.0)
```

### First-order SPPA at the Planning Subzone Level

**Geospatial data wrangling**

Extracting study area

```{r}
pg <- mpsz_cl |>
  filter(PLN_AREA_N == "PUNGGOL")
tm <- mpsz_cl |>
  filter(PLN_AREA_N == "TAMPINES")
ck <- mpsz_cl |>
  filter(PLN_AREA_N == "CHOA CHU KANG")
jw <- mpsz_cl |>
  filter(PLN_AREA_N == "JURONG WEST")
```

```{r, fig.width=10, fig.height=8}
par(mfrow=c(2,2))
plot(st_geometry(pg), main = "Ponggol")
plot(st_geometry(tm), main = "Tampines")
plot(st_geometry(ck), main = "Choa Chu Kang")
plot(st_geometry(jw), main = "Jurong West")
```

Creating ***owin*** object

```{r}
pg_owin = as.owin(pg)
tm_owin = as.owin(tm)
ck_owin = as.owin(ck)
jw_owin = as.owin(jw)
```

Combining point events object and owin object

```{r}
childcare_pg_ppp = childcare_ppp[pg_owin]
childcare_tm_ppp = childcare_ppp[tm_owin]
childcare_ck_ppp = childcare_ppp[ck_owin]
childcare_jw_ppp = childcare_ppp[jw_owin]
```

Trasnform the unit of measurement from metre to kilometre.

```{r}
childcare_pg_ppp.km = rescale.ppp(childcare_pg_ppp, 1000, "km")
childcare_tm_ppp.km = rescale.ppp(childcare_tm_ppp, 1000, "km")
childcare_ck_ppp.km = rescale.ppp(childcare_ck_ppp, 1000, "km")
childcare_jw_ppp.km = rescale.ppp(childcare_jw_ppp, 1000, "km")
```

Plot these four study areas and the locations of the childcare centres.

```{r, fig.width=10, fig.height=8}
par(mfrow=c(2,2))
plot(unmark(childcare_pg_ppp.km), 
     main="Punggol")
plot(unmark(childcare_tm_ppp.km), 
     main="Tampines")
plot(unmark(childcare_ck_ppp.km), 
     main="Choa Chu Kang")
plot(unmark(childcare_jw_ppp.km), 
     main="Jurong West")
```

**Clark-Evans test**

Choa Chu Kang planning area

```{r}
clarkevans.test(childcare_ck_ppp,
                correction="none",
                clipregion=NULL,
                alternative=c("two.sided"),
                nsim=999)
```

Tampines planning area

```{r}
clarkevans.test(childcare_tm_ppp,
                correction="none",
                clipregion=NULL,
                alternative=c("two.sided"),
                nsim=999)
```

Punggol planning area

```{r}
clarkevans.test(childcare_pg_ppp,
                correction="none",
                clipregion=NULL,
                alternative=c("two.sided"),
                nsim=999)
```

Jurong West planning area

```{r}
clarkevans.test(childcare_jw_ppp,
                correction="none",
                clipregion=NULL,
                alternative=c("two.sided"),
                nsim=999)
```

::: callout-important
**Statistical Conclusion**

- **Hypotheses:**

  - $H_0$: The spatial point distribution of childcare centres in these 4 planning areas follows Complete Spatial Randomness (CSR), where $R = 1$.

  - $H_1$: The spatial point distribution of childcare centres in Jurong West is not randomly distributed ($R \neq 1$).

- **Decision:** Since the p-value is substantially smaller than the significance threshold $\alpha = 0.05$, we reject the null hypothesis $H_0$.

- **Inference:** With an aggregation index of $R$ ($R < 1$), there is statistically significant evidence at the 95% confidence level to conclude that childcare centres in the Jurong West planning area exhibit a **spatially clustered** pattern rather than a random spatial distribution.
:::

**Computing KDE surfaces by planning area**

*bw.diggle* method is used to compute the KDE of these 4 planning areas

```{r, fig.width=10, fig.height=8}
par(mfrow=c(2,2))
plot(density(childcare_pg_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Punggol")
plot(density(childcare_tm_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Tempines")
plot(density(childcare_ck_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Choa Chu Kang")
plot(density(childcare_jw_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Jurong West")
```

## Part 2 Second-order Spatial Point pattern Analysis methods

Second-order spatial point pattern analysis examines the spatial relationships between points in a pattern, specifically focusing on how the presence of one point influences the location of others. It goes beyond simply describing the overall density of points (first-order effects) by investigating clustering, dispersion, or randomness at various spatial scales.

In this part, appropriate functions of spatstat are used to discover the spatial point processes of childcare centres in Singapore to answer the specific questions:

- are the children centres in Singapore randomly distributed throughout the country?
- if the answer is not, then the next logical question is where are the locations with higher concentration of childcare centres?

::: callout-note
## refereces

Methods used for second-order spatial point pattern analysis

+---------------------------+-------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------+
|                           |                                                                                                                               |                                                                       |
+===========================+===============================================================================================================================+=======================================================================+
| **Method / Function**     | **Purpose**                                                                                                                   | **Interpretation relative to CSR**                                    |
+---------------------------+-------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------+
| **G-Function**            | Measures the cumulative distribution of distances from an arbitrary event to its nearest event.                               | • **Above CSR:** Clustering                                           |
|                           |                                                                                                                               |                                                                       |
| \                         |                                                                                                                               | \                                                                     |
|                           |                                                                                                                               |                                                                       |
| `Gest()`                  |                                                                                                                               | • **Below CSR:** Regularity / Dispersion                              |
+---------------------------+-------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------+
| **F-Function**            | Measures the distribution of distances from an arbitrary point in space to the nearest observed event (empty space function). | • **Below CSR:** Clustering (larger empty spaces)                     |
|                           |                                                                                                                               |                                                                       |
| \                         |                                                                                                                               | \                                                                     |
|                           |                                                                                                                               |                                                                       |
| `Fest()`                  |                                                                                                                               | • **Above CSR:** Regularity / Dispersion                              |
+---------------------------+-------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------+
| **K-Function**            | Evaluates the expected number of events within distance $r$ of a typical event, normalized by intensity.                      | • **Above CSR:** Clustering at scale $r$                              |
|                           |                                                                                                                               |                                                                       |
| \                         |                                                                                                                               | \                                                                     |
|                           |                                                                                                                               |                                                                       |
| `Kest()`                  |                                                                                                                               | • **Below CSR:** Regularity / Dispersion                              |
+---------------------------+-------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------+
| **L-Function**            | Linearized, variance-stabilized transformation of Ripley's K-function, typically plotted as $L(r) - r$.                       | • $L(r) - r > 0$: Clustering                                          |
|                           |                                                                                                                               |                                                                       |
| \                         |                                                                                                                               | \                                                                     |
|                           |                                                                                                                               |                                                                       |
| `Lest()`                  |                                                                                                                               | • $L(r) - r < 0$: Regularity / Dispersion                             |
|                           |                                                                                                                               |                                                                       |
|                           |                                                                                                                               | \                                                                     |
|                           |                                                                                                                               |                                                                       |
|                           |                                                                                                                               | • $L(r) - r = 0$: Complete Spatial Randomness                         |
+---------------------------+-------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------+
| **Monte Carlo Envelopes** | Generates simulation envelopes (critical bands) under CSR to statistically test null hypotheses ($H_0$).                      | • **Outside Envelope:** Significant departure from CSR ($p < \alpha$) |
|                           |                                                                                                                               |                                                                       |
| \                         |                                                                                                                               | \                                                                     |
|                           |                                                                                                                               |                                                                       |
| `envelope()`              |                                                                                                                               | • **Inside Envelope:** Cannot reject CSR                              |
+---------------------------+-------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------+
:::

### Analysing Spatial Point Process Using G-Function

#### Choa Chu Kang planning area

Computing G-function estimation

```{r}
set.seed(1234)
```

Use *gest()* of **spatstat** package to compute G-function

```{r}
G_CK = Gest(childcare_ck_ppp, correction = "border")
plot(G_CK, xlim=c(0,500))
```

**Performing complete spatial randomness test**

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

Ho = The distribution of childcare services at Choa Chu Kang are randomly distributed.

H1= The distribution of childcare services at Choa Chu Kang are not randomly distributed.

The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

Monte Carlo test with G-fucntion

```{r}
G_CK.csr <- envelope(childcare_ck_ppp, Gest, nsim = 999)
```

```{r}
plot(G_CK.csr)
```

#### Tampines planning area

Computing G-function estimation

```{r}
G_tm = Gest(childcare_tm_ppp, correction = "best")
plot(G_tm)
```

Performing complete spatial randomness test

```{r}
G_tm.csr <- envelope(childcare_tm_ppp, Gest, correction = "all", nsim = 999)
```

```{r}
plot(G_tm.csr)
```

### Analysing spatial point process using F-Function

#### Choa Chu Kang planning area

Using *fest()* of **spatstat** package to compute F-function estimation

```{r}
F_CK = Fest(childcare_ck_ppp)
plot(F_CK)
```

**Performing Complete spatial randomness test**

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

Ho = The distribution of childcare services at Choa Chu Kang are randomly distributed.

H1= The distribution of childcare services at Choa Chu Kang are not randomly distributed.

The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

Monte Carlo test with F-fucntion

```{r}
F_CK.csr <- envelope(childcare_ck_ppp, Fest, nsim = 999)
```

```{r}
plot(F_CK.csr)
```

#### Tampines planning area

```{r}
F_tm = Fest(childcare_tm_ppp, correction = "best")
plot(F_tm)
```

```{r}
F_tm.csr <- envelope(childcare_tm_ppp, Fest, correction = "all", nsim = 999)
```

```{r}
plot(F_tm.csr)
```

### Analysing spatial point process using K-Function

#### Choa Chu Kang planning area

**Computing K-function estimate**

```{r}
K_ck = Kest(childcare_ck_ppp, correction = "Ripley")
plot(K_ck, . -r ~ r, ylab= "K(d)-r", xlab = "d(m)")
```

**Performing complete spatial randomness test**

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

Ho = The distribution of childcare services at Choa Chu Kang are randomly distributed.

H1= The distribution of childcare services at Choa Chu Kang are not randomly distributed.

The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

```{r}
K_ck.csr <- envelope(childcare_ck_ppp, Kest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
plot(K_ck.csr, . - r ~ r, xlab="d", ylab="K(d)-r")
```

#### Tampines planning area

```{r}
K_tm = Kest(childcare_tm_ppp, correction = "Ripley")
plot(K_tm, . -r ~ r, 
     ylab= "K(d)-r", xlab = "d(m)", 
     xlim=c(0,1000))
```

```{r}
K_tm.csr <- envelope(childcare_tm_ppp, Kest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
plot(K_tm.csr, . - r ~ r, 
     xlab="d", ylab="K(d)-r", xlim=c(0,500))
```

### Analysing spatial point process using L-Function

#### Choa Chu Kang planning area

Computing L-Function estimation

```{r}
L_ck = Lest(childcare_ck_ppp, correction = "Ripley")
plot(L_ck, . -r ~ r, 
     ylab= "L(d)-r", xlab = "d(m)")
```

Performing complete spatial randomness test

```{r}
L_ck.csr <- envelope(childcare_ck_ppp, Lest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
plot(L_ck.csr, . - r ~ r, xlab="d", ylab="L(d)-r")
```

#### Tampines planning area

```{r}
L_tm = Lest(childcare_tm_ppp, correction = "Ripley")
plot(L_tm, . -r ~ r, 
     ylab= "L(d)-r", xlab = "d(m)", 
     xlim=c(0,1000))
```

```{r}
L_tm.csr <- envelope(childcare_tm_ppp, Lest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
plot(L_tm.csr, . - r ~ r, 
     xlab="d", ylab="L(d)-r", xlim=c(0,500))
```

## Part 3 Learning more about test methods

### Clark-Evan test for Nearest Neighbour Analysis

The **Clark-Evans test** is a distance-based statistical method used in Nearest Neighbor Analysis (NNA) to determine whether a spatial point pattern exhibits **clustering**, **regularity (dispersion)**, or **Complete Spatial Randomness (CSR)**.

**Key Concept & Aggregation Index (**$R$)

The test compares the mean observed nearest neighbor distance ($\bar{r}_O$) to the expected mean distance under CSR ($\bar{r}_E$):

\$\$R = \\frac{\\bar{r}\_O}{\\bar{r}\_E}\$\$

- $R = 1$: The pattern is **randomly distributed** (consistent with CSR).

- $R < 1$: Observed points are closer together than expected $\rightarrow$ **Clustered pattern**.

- $R > 1$: Observed points are farther apart than expected $\rightarrow$ **Regular / Dispersed pattern** ($R \approx 2.15$ indicates a perfectly hexagonal lattice).

**Hypothesis Testing**

- **Null Hypothesis (**$H_0$): The point events are randomly distributed across the study region.

- **Alternative Hypothesis (**$H_1$): The point events are not randomly distributed (or specifically clustered: $R < 1$, or regular: $R > 1$).

**Implementation in R (`spatstat`)**

In R, the test is performed using `clarkevans.test()` from the [spatstat.explore](https://r4gdsa.netlify.app/chap04) package.

- **Standard Z-Test:**

  R

  ```         
  clarkevans.test(childcareSG_ppp,                  correction = "none",                  
  clipregion = "sg_owin",                  
  alternative = "clustered") 
  ```

- **Monte Carlo Simulation Test:**

  Compares the observed $R$ value against simulated realizations of CSR to derive an exact empirical $p$-value:

  R

  ```         
  clarkevans.test(childcareSG_ppp,                     correction = "none",                  
  clipregion = "sg_owin",                  
  alternative = "clustered",                  
  method = "MonteCarlo",                  
  nsim = 999) 
  ```

### Kernel Density Estimation (KDE) method

**Kernel Density Estimation (KDE)** is a fundamental first-order [Spatial Point Patterns Analysis](https://r4gdsa.netlify.app/chap04) (1st-SPPA) technique used within Exploratory Spatial Data Analysis (ESDA). It transforms discrete, observed point events into a continuous density surface, revealing spatial hotspots and variations in event intensity without imposing strict parametric assumptions.

**Core Mechanics in `spatstat`**

- **Basic Function:** Computed using `density()` on a planar point pattern (`ppp`) object.

- **Pixel Image Output:** The resulting object belongs to class `im`, representing a 2D regular pixel array where each cell contains the estimated point density.

- **Unit Rescaling:** Because default projected coordinates (such as SVY21 in meters) produce very small numbers (points per $\text{m}^2$), [`rescale.ppp()`](https://r4gdsa.netlify.app/chap04#cb33-1) is commonly applied to convert units to kilometers (points per $\text{km}^2$) for clearer interpretation.

**Key Tuning Parameters & Methods**

- **Bandwidth (**$\sigma$) Selection: Controls the degree of smoothing.

  \

  - `bw.diggle()`: Optimal for detecting isolated, tight clusters amidst random noise.

  - `bw.ppl()`: Recommended when the point pattern predominantly consists of tight clusters.

  - `bw.CvL()` & `bw.scott()`: Alternative cross-validation and rule-of-thumb bandwidth estimators.

- **Kernel Smoothing Types:** Defines the weighting function around each point (`gaussian` by default, `epanechnikov`, `quartic`, or `disc`).

- **Edge Correction:** Setting `edge = TRUE` adjusts for intensity underestimation near the study boundary (window border).

- **Fixed vs. Adaptive Bandwidth:**

  \

  - **Fixed (`density()`):** Uses a uniform bandwidth $\sigma$ across the entire region; sensitive to areas with extreme density variations (e.g., dense urban centers vs. sparse rural areas).

  - **Adaptive (`adaptive.density()`):** Dynamically adjusts the bandwidth based on local point concentration, applying smaller kernels in dense areas and wider kernels in sparse zones.

**Cartographic Visualization Pipeline**

To generate presentation-ready maps using packages like `tmap`:

1.  Convert the `spatstat` `im` object to a raster using `rast()` from the `terra` package.

2.  Assign the coordinate reference system (e.g., `crs(kde_raster) <- "EPSG:3414"`).

3.  Render continuous choropleth surfaces using `tm_shape()` + `tm_raster()`.

### Second-order: K-Function

**K-Function** is a multi-scale second-order [Spatial Point Patterns Analysis](https://r4gdsa.netlify.app/chap04) method that evaluates the spatial dependence (clustering, dispersion, or randomness) between point events across varying distance scales ($r$).

Unlike nearest-neighbor statistics (e.g., Clark-Evans or the G-function) which focus solely on the single closest point, the K-function evaluates the total number of points within a radius $r$ around every typical point.

**Mathematical Concept**

The theoretical K-function is defined as:

\$\$K(r) = \\frac{1}{\\lambda} E\[\\text{number of extra points within distance } r \\text{ of an arbitrary point}\]\$\$

- $\lambda$ represents the mean point intensity across the study region.

- Under Complete Spatial Randomness (CSR), the expected number of points within radius $r$ is $\lambda \pi r^2$, giving the theoretical baseline:

  \

  \$\$K\_{theo}(r) = \\pi r\^2\$\$

**Interpretation of Results**

- $K(r) > \pi r^2$: More points are observed within radius $r$ than expected by chance $\rightarrow$ **Clustering** at scale $r$.

- $K(r) < \pi r^2$: Fewer points are observed within radius $r$ than expected by chance $\rightarrow$ **Regularity / Dispersion** at scale $r$.

- $K(r) = \pi r^2$: Pattern aligns with Complete Spatial Randomness (CSR).

**Implementation in R (`spatstat`)**

Using the [`Kest()`](https://rdrr.io/cran/spatstat/man/Kest.html) and [`envelope()`](https://rdrr.io/cran/spatstat/man/envelope.html) functions from `spatstat`:

- **Computing the Empirical K-Function:**

  \

  R

  ```         
  # Estimate K with edge correction (e.g., Ripley/isotropic)
  K_est <- Kest(childcare_ck_ppp, correction = "Ripley")  

  # Plot K(r) centered against r (K(r) - r vs r) 
  plot(K_est, . - r ~ r, ylab = "K(d)-r", xlab = "d(m)") 
  ```

- **Testing Complete Spatial Randomness (Monte Carlo Simulation):**

  \

  R

  ```         
  # Generate Monte Carlo simulation envelopes 
  K_csr <- envelope(childcare_ck_ppp, Kest, nsim = 99, rank = 1, glocal = TRUE)  

  # Plot observed K against simulation envelopes 
  plot(K_csr, . - r ~ r, xlab = "d", ylab = "K(d)-r") 
  ```

  - If the observed curve lies **above the upper envelope**, clustering is statistically significant ($p < \alpha$).

  - If the curve lies **below the lower envelope**, dispersion/regularity is statistically significant.

### Second-order: G-Function

The **G-Function** (Nearest Neighbor Distance Distribution Function) is a second-order [Spatial Point Patterns Analysis](https://r4gdsa.netlify.app/chap04) method that measures the cumulative distribution of distances from an arbitrary event to its nearest neighboring event.

**Key Mathematical Concept**

The function estimates the probability $G(r)$ that an event's nearest neighbor is located within a radial distance $r$:

\$\$G(r) = P(d_i \\le r)\$\$

Under the null hypothesis of Complete Spatial Randomness (CSR), the theoretical distribution $G_{theo}(r)$ for a homogeneous Poisson process with intensity $\lambda$ is:

\$\$G\_{theo}(r) = 1 - \\exp(-\\lambda \\pi r\^2)\$\$

**Interpretation Relative to CSR**

- $G(r) > G_{theo}(r)$: The curve rises faster than expected at short distances, indicating that events are located closer to one another $\rightarrow$ **Clustering**.

- $G(r) < G_{theo}(r)$: The curve rises slower than expected, indicating events maintain distance from one another $\rightarrow$ **Regularity / Dispersion**.

- $G(r) \approx G_{theo}(r)$: The point pattern conforms to **Complete Spatial Randomness (CSR)**.

**Implementation in R (`spatstat`)**

- **Estimation with Edge Correction:**

  Edge corrections (such as `"border"`, `"hanisch"`, or `"best"`) adjust for unobserved points beyond the study boundary:

  \

  R

  ```         
  # Compute G-function estimation 
  G_est <- Gest(childcare_ck_ppp, correction = "border")  

  # Plot empirical vs theoretical G(r) 
  plot(G_est, xlim = c(0, 500)) 
  ```

- **Complete Spatial Randomness (CSR) Hypothesis Testing:**

  Uses [envelope()](https://rdrr.io/cran/spatstat/man/envelope.html) to generate Monte Carlo simulation critical bands:

  \

  R

  ```         
  # Monte Carlo simulation test under CSR (e.g., 999 simulations) 
  G_csr <- envelope(childcare_ck_ppp, Gest, correction = "all", nsim = 999)  

  # Plot observed G(r) against simulation envelope 
  plot(G_csr) 
  ```

  - **Above Upper Envelope:** Statistically significant clustering at scale $r$ ($p < \alpha$).

  - **Below Lower Envelope:** Statistically significant regularity/dispersion at scale $r$ ($p < \alpha$).

### Second-order: F-Function

The **F-Function** (also known as the **Empty Space Function** or point-to-event distance distribution) is a second-order [Spatial Point Patterns Analysis](https://r4gdsa.netlify.app/chap05) method that measures the distribution of distances from an arbitrary location in the study area to the nearest observed event.

Unlike the G-function (which measures distances between event pairs), the F-function samples the empty space across the entire study window.

**Mathematical Concept**

The function estimates the cumulative probability $F(r)$ that an arbitrary point chosen at random within the observation window has at least one event within radial distance $r$:

\$\$F(r) = P(d(x, Y) \\le r)\$\$

where $d(x, Y)$ is the shortest distance from an arbitrary point $x$ to the nearest event in point pattern $Y$.

Under Complete Spatial Randomness (CSR) for a homogeneous Poisson process with intensity $\lambda$, the theoretical baseline is:

\$\$F\_{theo}(r) = 1 - \\exp(-\\lambda \\pi r\^2)\$\$

**Interpretation Relative to CSR**

Because the F-function measures empty spaces:

- $F(r) < F_{theo}(r)$: The curve rises slower than CSR, meaning empty spaces are larger than expected by chance $\rightarrow$ **Clustering** (events group together, leaving wide gaps).

- $F(r) > F_{theo}(r)$: The curve rises faster than CSR, meaning points are evenly spread throughout the space $\rightarrow$ **Regularity / Dispersion**.

- $F(r) \approx F_{theo}(r)$: The pattern conforms to **Complete Spatial Randomness (CSR)**.

**Implementation in R (`spatstat`)**

Using the [`Fest()`](https://rdrr.io/cran/spatstat/man/Fest.html) and [`envelope()`](https://rdrr.io/cran/spatstat/man/envelope.html) functions from `spatstat`:

- **Computing the Empirical F-Function:**

  \

  R

  ```         
  # Estimate F with edge correction (e.g., "best" or "raw") 
  F_ck <- Fest(childcare_ck_ppp, correction = "best")  

  # Plot empirical vs theoretical F(r) 
  plot(F_ck) 
  ```

- **Complete Spatial Randomness (CSR) Hypothesis Testing:**

  \

  R

  ```         
  # Monte Carlo simulation test under CSR (e.g., 999 simulations) 
  F_ck_csr <- envelope(childcare_ck_ppp, Fest, nsim = 999)  

  # Plot observed F(r) against simulation envelopes 
  plot(F_ck_csr) 
  ```

  - **Below Lower Envelope:** Statistically significant clustering ($p < \alpha$).

  - **Above Upper Envelope:** Statistically significant regularity/dispersion ($p < \alpha$).

### Second-order: K-Function

The **K-Function** (Ripley's K-Function) is a multi-scale second-order [Spatial Point Patterns Analysis](https://r4gdsa.netlify.app/chap05) method used to evaluate whether point events exhibit clustering, dispersion, or randomness across multiple distance scales ($r$).

Unlike nearest-neighbor statistics, which only consider the distance to the single closest point, the K-function evaluates the total number of points found within a given radius $r$ of any arbitrary point.

**Mathematical Concept**

The theoretical K-function is defined as:

\$\$K(r) = \\frac{1}{\\lambda} \\mathbb{E}\[\\text{number of additional events within distance } r \\text{ of a typical event}\]\$\$

- $\lambda$ is the overall intensity of the point pattern.

- Under Complete Spatial Randomness (CSR) for a homogeneous Poisson process, the expected number of events within radius $r$ is $\lambda \pi r^2$, yielding the theoretical baseline:

  \

  \$\$K\_{theo}(r) = \\pi r\^2\$\$

**Interpretation Relative to CSR**

- $K(r) > \pi r^2$: More events are observed within distance $r$ than expected under randomness $\rightarrow$ **Clustering** at scale $r$.

- $K(r) < \pi r^2$: Fewer events are observed within distance $r$ than expected $\rightarrow$ **Regularity / Dispersion** at scale $r$.

- $K(r) = \pi r^2$: Follows Complete Spatial Randomness (CSR).

**Implementation in R (`spatstat`)**

- **Computing Empirical K-Function:**

  Estimated using [`Kest()`](https://rdrr.io/cran/spatstat/man/Kest.html) with an edge-correction method (such as Ripley's isotropic correction):

  \

  R

  ```         
  # Estimate K-function 
  K_ck <- Kest(childcare_ck_ppp, correction = "Ripley")  

  # Plot empirical vs theoretical K(r) centered against r (K(r) - r vs r) 
  plot(K_ck, . - r ~ r, ylab = "K(d)-r", xlab = "d(m)") 
  ```

- **Complete Spatial Randomness (CSR) Hypothesis Testing:**

  Evaluated using [`envelope()`](https://rdrr.io/cran/spatstat/man/envelope.html) to run Monte Carlo simulations and establish critical confidence bands:

  \

  R

  ```         
  # Run Monte Carlo simulation envelopes under CSR 
  K_ck.csr <- envelope(childcare_ck_ppp, Kest, nsim = 99, rank = 1, glocal = TRUE)  

  # Plot observed K against simulation envelopes 
  plot(K_ck.csr, . - r ~ r, xlab = "d", ylab = "K(d)-r") 
  ```

  - **Above Upper Envelope:** Statistically significant clustering at scale $r$ ($p < \alpha$).

  - **Below Lower Envelope:** Statistically significant regularity/dispersion at scale $r$ ($p < \alpha$).

### Second-order: L-Function

The **L-Function** (Besag’s L-Function) is a variance-stabilized, linearized transformation of [Ripley’s K-Function](https://r4gdsa.netlify.app/chap05#analysing-spatial-point-process-using-k-function) used in second-order [Spatial Point Patterns Analysis](https://r4gdsa.netlify.app/chap05).

While the K-function grows quadratically with distance ($K(r) \approx \pi r^2$ under Complete Spatial Randomness), the L-function transforms this into a linear relationship, making departures from randomness easier to visually inspect and interpret across distance scales ($r$).

**Mathematical Concept**

Besag’s transformation scales the K-function as:

\$\$L(r) = \\sqrt{\\frac{K(r)}{\\pi}}\$\$

Under Complete Spatial Randomness (CSR), $K_{theo}(r) = \pi r^2$, which simplifies to:

\$\$L\_{theo}(r) = r \\quad \\iff \\quad L(r) - r = 0\$\$

In practice, the L-function is almost always analyzed and plotted as the centered function $L(r) - r$ (or $L(d) - d$) against distance $r$.

**Interpretation Relative to CSR**

- $L(r) - r > 0$: Observed spatial density within radius $r$ exceeds CSR expectation $\rightarrow$ **Clustering** at scale $r$.

- $L(r) - r < 0$: Observed spatial density within radius $r$ is less than CSR expectation $\rightarrow$ **Regularity / Dispersion** at scale $r$.

- $L(r) - r = 0$: Conforms to Complete Spatial Randomness (CSR).

**Implementation in R (`spatstat`)**

Using [`Lest()`](https://rdrr.io/cran/spatstat/man/Lest.html) and [`envelope()`](https://rdrr.io/cran/spatstat/man/envelope.html) from `spatstat`:

- **Computing Empirical L-Function:**

  \

  R

  ```         
  # Estimate L-function with edge correction (e.g., Ripley/isotropic) 
  L_ck <- Lest(childcare_ck_ppp, correction = "Ripley")  

  # Plot centered L(r) - r against distance r 
  plot(L_ck, . - r ~ r, ylab = "L(d)-r", xlab = "d(m)") 
  ```

- **Complete Spatial Randomness (CSR) Hypothesis Testing:**

  \

  R

  ```         
  # Generate Monte Carlo simulation envelopes under CSR 
  L_ck_csr <- envelope(childcare_ck_ppp, Lest, nsim = 99, rank = 1, glocal = TRUE)  

  # Plot observed L(r) - r against the simulation envelope 
  plot(L_ck_csr, . - r ~ r, xlab = "d", ylab = "L(d)-r") 
  ```

  - **Above Upper Envelope:** Statistically significant clustering at distance $r$ ($p < \alpha$).

  - **Below Lower Envelope:** Statistically significant regularity/dispersion at distance $r$ ($p < \alpha$).

  - **Inside Envelopes:** No statistically significant departure from CSR.
