---
title: "In-class_Ex05"
author: "Hung Thai"
date: "Sep 26, 2026"
date-modified: "last-modified"
execute:
  echo: true
  eval: true
  mesage: false
  warning: false
  freeze: true
format: 
  html:
    toc: true
    toc-depth: 4
    code-copy: true
editor: visual
---

## Loading the packages

```{r}
pacman::p_load(sf, tmap, tidyverse, sfdep)
```

## Importing Hunan shapefile and GDP table

Import Hunan shapefile

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```

Import Hunan GDP file:

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

Join Hunan & Hunan2012 data.frames and tidy the data table by retaining NAME_2, ID_3, NAME_3, County, GDPPC, GIO, Agri and Service

```{r}
hunan_GDPPC_ <- left_join(hunan,hunan2012)
```

```{r}
hunan_GDPPC <- hunan_GDPPC_ %>%
  select(1:3, 7, 15, 16, 33)
```

## Global mearuse of Spatial Association

**Step1: deriving Queen's contiguity weights: sfdep methods**

```{r}
wm_q <- hunan_GDPPC %>%
  mutate (nb= st_contiguity(geometry),
          wt= st_weights(nb, style = "W"),
          .before=1)

```

```{r}
names(wm_q)
```

**Step2: computing Global Moran's I**

```{r}
moranI <- global_moran(wm_q$GDPPC,
                       wm_q$nb,
                       wm_q$wt)

glimpse(moranI)
```

**Computing Local Moran's I**

```{r}
lisa <- wm_q %>%
  mutate(local_moran = local_moran(
    GDPPC, nb, wt, nsim = 99),
    .before =1) %>%
  unnest(local_moran)
```

```{r}
lisa_sig <- lisa %>%
  filter(p_ii_sim < 0.05)
tmap_mode("plot")
tm_shape(lisa) +
  tm_polygons() +
  tm_borders(fill_alpha=0.5) +
tm_shape(lisa_sig) + 
  tm_polygons(fill = "mean",
              fill.legend = tm_legend(
                text.size = 1.15,
                width =6
              )) +
  tm_borders(fill_alpha = 0.4) +
  tm_layout(legend.title.size = 1.25)

```
