---
title: "In-class_Ex04"
author: "Hung Thai"
date: "Sep 19, 2026"
date-modified: "last-modified"
execute:
  echo: true
  eval: true
  mesage: false
  warning: false
  freeze: true
format: 
  html:
    toc: true
    toc-depth: 4
    code-copy: true
editor: visual
---

## Loading the packages

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse, knitr, GWmodel, ggstatsplot)
```

## Importing Hunan shapefile and GDP table

Import Hunan shapefile

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```

Import Hunan GDP file:

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

Join Hunan & Hunan2012 data.frames and tidy the data table by retaining NAME_2, ID_3, NAME_3, County, GDPPC, GIO, Agri and Service

```{r}
hunan <- left_join(hunan,hunan2012) %>%
  select(1:3, 7, 15, 16, 31, 32)
```

## Plot choropleth map

```{r}
#| fig-width: 16
#| fig-height: 6


basemap <- tm_shape(hunan) +
  tm_polygons() +
  tm_text("NAME_3", size=0.5)

gdppc <- qtm(hunan, "GDPPC") +
  tm_layout(
    legend.position = c("left", "bottom"),
    legend.text.size = 0.6,
    legend.title.size = 0.7)
tmap_arrange(basemap, gdppc, asp=1, ncol=2)
```

**Converting to SpatialPolygonDataFrame**

```{r}
hunan_sp <- hunan %>%
  as_Spatial()
```

## Geographically weighted summary Statistics (GSWW) with adaptive bandwidth

Cross validation

```{r}
bw_CV <- bw.gwr(GDPPC ~ 1,
                data = hunan_sp,
                approach = "CV",
                adaptive = TRUE,
                kernel = "bisquare",
                longlat = T)
```

Cross validation with adaptive = FALSE

```{r}
bw_CVf <- bw.gwr(GDPPC ~ 1,
                data = hunan_sp,
                approach = "CV",
                adaptive = FALSE,
                kernel = "bisquare",
                longlat = T)
```

AIC correcting

```{r}
bw_AIC <- bw.gwr(GDPPC ~ 1,
                data = hunan_sp,
                approach = "AICc",
                adaptive = TRUE,
                kernel = "bisquare",
                longlat = T)
```

## Computing geographically weighted summary statistic

```{r}
gwstat <- gwss(data = hunan_sp,
               vars = "GDPPC",
               bw = bw_AIC,
               kernel = "bisquare",
               quantile = TRUE,
               adaptive = TRUE,
               longlat = T)
```

GWSTAT:

```{r}
gwstat
```

## Preparing the output data

Extract SDF data table from gwss object output from gwss() for subsequent geovilisation.

```{r}
gwstat_df <- as.data.frame(gwstat$SDF)
```

Next cbind() is used to append the newly derived data.frame onto hunan sf data frame.

```{r}
hunan_gstat <- cbind(hunan, gwstat_df)
```

## Visualising geographically weighted summary statistics

```{r}
tm_shape(hunan_gstat) +
  tm_polygons(fill = "GDPPC_LM",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n=5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "local Mean")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of geographically weighted mean") +
  tm_layout(legend.position = c("left", "bottom"))

```

## Geographically Weighted Correlation

Relationship between GDP per capita and Gross Industry Output

Conventional statistical solution

```{r}
ggscatterstats(
  data = hunan2012, 
  x = Agri, 
  y = GDPPC,
  xlab = "Gross Agriculture Output", 
  ylab = "GDP per capita", 
  label.var = County, 
  label.expression = Agri > 10000 & GDPPC > 50000, 
  point.label.args = list(alpha = 0.7, size = 4, color = "grey50"),
  xfill = "#CC79A7", 
  yfill = "#009E73", 
  title = "Relationship between GDP PC and Gross Agriculture Output")
```

GeoVisualisation

```{r}
#| fig-width: 12
#| fig-height: 8
#| 
gdppcmap <- tm_shape(hunan) + 
  tm_polygons(
    fill = "GDPPC",
    fill.scale = tm_scale_intervals(
      n = 5,
      style = "quantile",
      values = "brewer.oranges")) +
  tm_borders(fill_alpha = 0.25,
             lwd = 0.5) +
  tm_title("Distribution of GDP per capita") +
  tm_layout(legend.position = c("left", "bottom")) 

giomap <- tm_shape(hunan) + 
  tm_polygons(
    fill = "GIO",
    fill.scale = tm_scale_intervals(
      n = 5,
      style = "quantile",
      values = "blue")) +
  tm_borders(fill_alpha = 0.25,
             lwd = 0.5) +
  tm_title("Distribution of Gross Industry Output") +
  tm_layout(legend.position = c("left", "bottom"))

tmap_arrange(gdppcmap, giomap, asp=1, ncol=2)
```

## Geospatial analytics solution

Determine the bandwidth

```{r}
bw <- bw.gwr(GDPPC ~ GIO, 
             data = hunan_sp, 
             approach = "AICc",
             kernel = "bisquare",
             adaptive = TRUE)
```

Computing gwCorrelation

```{r}
gwstats <- gwss(hunan_sp, 
                vars = c("GDPPC", "GIO"), 
                bw = bw,
                kernel = "bisquare",
                adaptive = TRUE, 
                longlat = T)
```

Extracting the SDF data table from gwss object then convert into data.frame.

```{r}
gwstat_df <- as.data.frame(gwstats$SDF) %>%
  select(c(12,13)) %>%
  rename(gwCorr = Corr_GDPPC.GIO,
         gwSpearman = Spearman_rho_GDPPC.GIO)
```

cbind() is used to append the newly derived data.frame onto hunan sf data.frame.

```{r}
hunan_Corr <- cbind(hunan, gwstat_df)
```

## Geospatial analytics solution

Determine the bandwidth

```{r}
bw <- bw.gwr(GDPPC ~ GIO, 
             data = hunan_sp, 
             approach = "AICc",
             kernel = "bisquare",
             adaptive = TRUE)
```

Computing gwCorrelation

```{r}
gwstats <- gwss(hunan_sp, 
                vars = c("GDPPC", "GIO"), 
                bw = bw,
                kernel = "bisquare",
                adaptive = TRUE, 
                longlat = T)
```

Extracting the SDF data table from gwss object

```{r}
gwstat_df <- as.data.frame(gwstats$SDF) %>%
  select(c(12,13)) %>%
  rename(gwCorr = Corr_GDPPC.GIO,
         gwSpearman = Spearman_rho_GDPPC.GIO)
```

Append the newly derived data.frame onto hunan sf data.frame.

```{r}
hunan_Corr <- cbind(hunan, gwstat_df)
```

## Visualising local correlation

Local correlation coefficient

```{r}
#| fig-width: 10
#| fig-height: 8
tm_shape(hunan_Corr) +
  tm_polygons(fill = "gwCorr",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "local Correlation Coefficient")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Local Spearman Rho", 
           size = 2.0) +
  tm_layout(frame = TRUE)
```

Local Spearman Coefficient

```{r}
#| fig-width: 10
#| fig-height: 8
tm_shape(hunan_Corr) +
  tm_polygons(fill = "gwSpearman",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "local Spearman Rho")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Local Spearman Rho", 
           size = 2.0) +
  tm_layout(frame = TRUE)
```

## References

Brunsdon, C. et. al. (2002) “Geographically weighted summary statistics - a framework for localised exploratory data analysis”, Computer, *Environment and Urban Systems*, Vol 26, pp. 501-525. Available as e-journal, SMU library.

Harris, P. & Brunsdon, C. (2010) “Exploring spatial variation and spatial relationships in freshwater acidification critical load data set for Great Britain using geographically weighted summary statistics”, *Computers & Geosciences*, Vol. 36, pp. 54-70. Available as e-journal, SMU library.

[ISSS626-ay2026-27aug In-class Exercise04](https://isss626-ay2026-27aug.netlify.app/in-class_ex/in-class_ex04/in-class_ex04)
